### 2019-12-13 doc/2020-CAV/paper.tex
Moved CAV paper to doc folder

\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{xtab}
\usepackage{stmaryrd}
\usepackage{amsmath, amssymb}
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

% OUR MACROS
\input{macros}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\title{End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
%
\titlerunning{End-to-End Verification of GR(1) Designs in SPARK}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Princeton University, Princeton NJ 08544, USA \and
Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{lncs@springer.com}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 150--250 words.
Manually designing control logic for reactive systems is time-consuming and error-prone. 
An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
Recently, there has been interest in synthesizing controllers from Generalized Reactivity(1) or GR(1) specifications, 
since computational complexity is relatively low, and several tools now exist. 
However, though these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
errors in tool implementation can lead to errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
Toward this end we have modified Salty, a tool that produces executable software implementations of controllers 
from GR(1) specifications in a variety of programming languages, to produce implementations in SPARK.
SPARK is both a language and associated set of verification tools, and so its use enables the desired ``end-to-end'' verification of a 
subset of properties comprising GR(1) specifications, in this case system initial and transition properties. 
In this paper, we discuss how we encode synthesized controllers and the necessary contracts and assertions needed for 
SPARK to verify the aforementioned properties automatically, i.e. without requiring additional annotations from the user. 
Since GR(1) specifications encode assumptions about the environment in which the controlled system operates, 
we also discuss how we handle these assumptions in the implementation, since there are several reasonable choices for how they could be handled. 
Finally, we discuss possible approaches for verifying the full set of GR(1) properties, i.e. including liveness.

\keywords{Reactive synthesis \and end-to-end verification \and functional verification.}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUCTION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Reactive systems must be capable of correctly responding to various inputs, 
e.g. originating from human users or events in the system's operational environment. 
The process of manually designing the control logic for such systems is both time-consuming and error prone. 
An alternative is to use ``correct-by-construction'' synthesis approaches to automatically generate 
the design of a system's control logic directly from specifications, which can 
reduce both design time and the likelihood of errors 
\cite{alur2016compositional}, \cite{fainekos2009temporal}, \cite{guo2014cooperative}, \cite{kupermann2001synthesizing}. 
In general, \emph{synthesis} is the process of automatically generating a design from a specification. 
More specifically, \emph{reactive synthesis} approaches generate designs in the context of an uncontrolled environment, 
assumptions about which are encoded in the specification.
Reactive synthesis approaches tend to have high computational complexity,  
so there is particular interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
the complexity of which is only polynomial in the size of the game graph encoded by the specification \cite{bloem2012}.
Synthesis from GR(1) specifications has been used to generate digital circuits \cite{ehlers2012symbolically} 
and controllers for aircraft power distribution \cite{xu2012case}, 
software-defined networks \cite{wang2013automated}, ground robots \cite{kress2007s},
and teams of unmanned vehicles \cite{apker2016}, to name a few.

A GR(1) specification $\varphi$ takes the form $\varphi = \varphi^e \limplies \varphi^s$, 
where $\varphi^e$ encodes assumptions about the \emph{environment} in which a system is to operate,
and $\varphi^s$ encodes guarantees the \emph{system} should make under those assumptions \cite{Ehlers2016}. 
More specifically, $\varphi$ takes the form 
$\varphi = (\varphi^e_i \land \varphi^e_t \land \varphi^e_l)  \limplies (\varphi^s_i \land \varphi^s_t \land \varphi^s_l)$,
where $\varphi^e_i$ and $\varphi^s_i$ are \emph{initial} properties, 
$\varphi^e_t$ and $\varphi^s_t$ are \emph{transition} or safety properties, and 
$\varphi^e_l$ and $\varphi^s_l$ are \emph{liveness} properties. 
For inputs in the set $\inputs$ controlled by the environment 
and outputs in the set $\outputs$ produced by the system, terms are:

\vspace{0.5em}

\noindent \begin{xtabular}{p{.045\columnwidth}p{.015\columnwidth}p{.860\columnwidth}}
 $\varphi^e_i$, $\varphi^s_i$ & - & Boolean formulas over $\inputs$ and $\outputs$, respectively, that characterize the initial state of the environment and system.
\end{xtabular}

\noindent \begin{xtabular}{p{.045\columnwidth}p{.015\columnwidth}p{.860\columnwidth}}
 $\varphi^e_t$, $\varphi^s_t$ & - &  Formulas of the form $\bigwedge_{j \in J} \always B_j$, where each $B_j$ is a Boolean combination of variables from $\inputs \cup \outputs$ and expressions of the form $\lnext v$, where $v \in \inputs$ for $\varphi^e_t$ and $v \in \inputs \cup \outputs$ for $\varphi^s_t$. These encode properties that should always hold as well as rules for how inputs and outputs are allowed to change based on current input and output values.
\end{xtabular}

\noindent \begin{xtabular}{p{.045\columnwidth}p{.015\columnwidth}p{.860\columnwidth}}
$\varphi^e_l$, $\varphi^s_l$ & - & Formulas of the form $\bigwedge_{j \in J} \always \eventually B_j$, where each $B_j$ is a Boolean formula over $\inputs \cup \outputs$. These encode properties that should hold infinitely often.
\end{xtabular}

\vspace{0.5em}

\noindent It is assumed that at each time step, the environment chooses an input from $\inputs$, and then the system chooses an output from $\outputs$ in response.

Several tools for synthesis from GR(1) specifications are available. 
For example, RATSY \cite{bloem2010ratsy} has a focus on circuit design and can synthesize designs encoded in 
BLIF (Berkeley Logic Interchange Format), Verilog, and HIF (HDL Intermediate Format). 
Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
LTLMoP \cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
and it synthesizes designs as hybrid controllers along with handler modules to help connect controllers to simulated or real-world systems.
Slugs \cite{Ehlers2016} is architected to allow users to easily tailor the synthesis algorithm for specific applications,
 e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and produces mathematical representations of controller designs.
Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
a back-end that turns controller designs into executable sofware implementations in a variety of programming languages.

Though all of these tools implement synthesis algorithms that are theoretically ``correct-by-construction,'' 
tool implementation errors could result in errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications. 
Toward this end, we have extended the Salty tool to produce software implementations of synthesized controller designs in SPARK. 
SPARK is both a programming language with a specification language and associated verification toolset \cite{hoang2015spark}. 
Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers through 
annotations such as assertions and loop invariants in the code. 
However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their functional specifications in SPARK. 
To date, we have largely achieved this goal for a subset of properties comprising GR(1) specifiations, 
i.e. system initial and transition properties, for moderately-sized controllers.
In what follows, in \sectref{sec:implementation} we discuss the SPARK implementation. 
In \sectref{sec:caseStudies}, we demonstrate the efficacy of our approach on examples pulled from a variety of sources.
In \sectref{sec:discussion}, we discuss possible approaches for proving liveness properties as well as design choice alternatives, 
e.g. for dealing with environment specifications.
We end with concluding remarks in

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% IMPLEMENTATION and Verification in SPARK
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementation and Verification in SPARK}
\label{sec:implementation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CASE STUDIES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Case Studies}
\label{sec:caseStudies}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CASE STUDIES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONCLUSIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{bibfile}

\end{document}

### 2020-01-20 doc/2020-CAV/paper.tex
Draft paper updates

@@
 \usepackage{amsmath, amssymb}
+\usepackage{listings}
 % If you use the hyperref package, please uncomment the following line
@@
 \begin{document}
+\lstset{language=Ada}
+\lstset{basicstyle=\ttfamily}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %
-\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
+\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
+James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
+Joffrey Heugenot\inst{3}\orcidID{2222--3333-4444-5555}}
 %
-\authorrunning{F. Author et al.}
+\authorrunning{L. Humphrey et al.}
 % First names are abbreviated in the running head.
@@
 %
-\institute{Princeton University, Princeton NJ 08544, USA \and
-Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-\email{lncs@springer.com}\\
-\url{http://www.springer.com/gp/computer-science/lncs} \and
-ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-\email{\{abc,lncs\}@uni-heidelberg.de}}
+\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
+\email{laura.humphrey@us.af.mil} \and
+LinQuest Corp., Beavercreek, OH 45431, USA \\
+\email{james.hamil@ctr.us.af.mil}  \and
+AdaCore, F-75009 Paris, France \\
+ \email{huguet@adacore.com}}
 %
@@
 An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
-Recently, there has been interest in synthesizing controllers from Generalized Reactivity(1) or GR(1) specifications, 
-since computational complexity is relatively low, and several tools now exist. 
-However, though these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
-errors in tool implementation can lead to errors in synthesized controllers. 
+Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
+since the required computational complexity is relatively low, and several tools now exist for GR(1) synthesis.
+However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
+errors in tool implementation can still lead to errors in synthesized controllers. 
 We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
-Toward this end we have modified Salty, a tool that produces executable software implementations of controllers 
-from GR(1) specifications in a variety of programming languages, to produce implementations in SPARK.
-SPARK is both a language and associated set of verification tools, and so its use enables the desired ``end-to-end'' verification of a 
-subset of properties comprising GR(1) specifications, in this case system initial and transition properties. 
-In this paper, we discuss how we encode synthesized controllers and the necessary contracts and assertions needed for 
-SPARK to verify the aforementioned properties automatically, i.e. without requiring additional annotations from the user. 
-Since GR(1) specifications encode assumptions about the environment in which the controlled system operates, 
-we also discuss how we handle these assumptions in the implementation, since there are several reasonable choices for how they could be handled. 
-Finally, we discuss possible approaches for verifying the full set of GR(1) properties, i.e. including liveness.
+Toward this end we have modified Salty -- a tool that produces executable software implementations of controllers 
+from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
+SPARK is both a language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.  
+In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
+a subset of properties comprising their original GR(1) specifications, namely system initial and transition properties. 
+We also discuss leasons learned about how to best encode controllers in SPARK for verification, 
+examples in which verification found unexpected controller behaviors, and 
+caveats related to the interpretation of GR(1) specifications.
+
+%In this paper, we discuss our experience to date using SPARK to implement controllers and to verify them against 
+%a subset of properties comprising their original GR(1) specifications, in this case system initial and transition properties. 
+%There are many ways to encode synthesized controllers in SPARK, so we describe the encoding that we found to be most efficient for verification. 
+%We also describe the contracts and assertions that we needed to generate within the controller implementation so that 
+%SPARK could verify the aforementioned properties automatically, i.e. without requiring additional annotations from the user.
+%We give some preliminary results on the approximate size of controllers that can be verified and the amount of time needed for verification, 
+%using a database of GR(1) specification examples pulled from a variety of sources. 
+%Finally, we close with a discussion of certain implementation choices that arise due to environment assumptions encoded in GR(1) specifications, 
+%some of which we have handled and some of which we have currently deferred. 
+%We also discuss possible approaches for verifying the full set of GR(1) properties, i.e. including liveness.
 
@@
 Synthesis from GR(1) specifications has been used to generate digital circuits \cite{ehlers2012symbolically} 
-and controllers for aircraft power distribution \cite{xu2012case}, 
-software-defined networks \cite{wang2013automated}, ground robots \cite{kress2007s},
-and teams of unmanned vehicles \cite{apker2016}, to name a few.
+and controllers for teams of unmanned vehicles \cite{apker2016}, ground robots \cite{kress2007s}, 
+software-defined networks \cite{wang2013automated}, and aircraft power distribution systems \cite{xu2012case}, to name a few.
 
@@
 
-\noindent It is assumed that at each time step, the environment chooses an input from $\inputs$, and then the system chooses an output from $\outputs$ in response.
+\noindent It is assumed that at each time step, the environment chooses an input from $\inputs$, then the system chooses an output from $\outputs$ in response. 
+The result of synthesis is a \emph{control protocol} or \emph{strategy} that can be expressed as a Moore machine, 
+with each state encoding the most recent input values from the environment and the most recent output values produced in response by the system.
 
-Several tools for synthesis from GR(1) specifications are available. 
-For example, RATSY \cite{bloem2010ratsy} has a focus on circuit design and can synthesize designs encoded in 
-BLIF (Berkeley Logic Interchange Format), Verilog, and HIF (HDL Intermediate Format). 
+Several tools for GR(1) synthesis are available. 
+For example, RATSY %(Requirements Analysis Tool with Synthesis) 
+\cite{bloem2010ratsy} has a focus on circuit design and can synthesize designs 
+encoded in BLIF, %(Berkeley Logic Interchange Format),
+Verilog, and HIF. %(HDL Intermediate Format). 
 Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
-LTLMoP \cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
-and it synthesizes designs as hybrid controllers along with handler modules to help connect controllers to simulated or real-world systems.
-Slugs \cite{Ehlers2016} is architected to allow users to easily tailor the synthesis algorithm for specific applications,
- e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and produces mathematical representations of controller designs.
+LTLMoP 
+%(Linear Temporal Logic MissiOn Planning)  
+\cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
+and it synthesizes designs as hybrid controllers with handler modules to help connect controllers to simulated or real-world systems. 
+TuLiP 
+%(Temporal Logic Planning) 
+\cite{TuLiP2011} has a similar focus and can synthesize controller implementations in Python. 
+Slugs 
+%(SmalL bUt Complete GROne Synthesizer) 
+\cite{Ehlers2016} is architected to allow users to easily tailor synthesis algorithms for specific applications,
+ e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
 Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
@@
 
-Though all of these tools implement synthesis algorithms that are theoretically ``correct-by-construction,'' 
-tool implementation errors could result in errors in synthesized controllers. 
+Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,'' 
+tool implementation errors could still result in errors in synthesized controllers. 
 We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications. 
-Toward this end, we have extended the Salty tool to produce software implementations of synthesized controller designs in SPARK. 
+Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK. 
 SPARK is both a programming language with a specification language and associated verification toolset \cite{hoang2015spark}. 
-Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers through 
-annotations such as assertions and loop invariants in the code. 
+Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers by
+annotating the code, e.g. with assertions and loop invariants. 
 However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
@@
 e.g. for dealing with environment specifications.
-We end with concluding remarks in
+We end with concluding remarks in \sectref{sec:conclusions}.
 
@@
 
+To explain the structure of the controllers we generate in SPARK, consider a simple example of a controller for a traffic light. 
+Color changes are triggered in every state in which an input signal's value is true.
+Let this input be called $tick$.
+The traffic light's color changes from red to green to yellow and back to red, repeating the cycle infinitely. 
+Let the output variables then be $red$, $yellow$, and $green$.
+The environment specifications are $\varphi^e_i = \top$, $\varphi^e_t = \top$, $\varphi^e_l = \always \eventually tick$, i.e.  
+$tick$ can be true or false in the initial state and has no constraints on how it transitions from state to state, but it must be true infinitely often.
+The system initial and liveness specifications are $\varphi^s_i = red \land \lnot yellow \land \lnot green$ and $\varphi^s_l =  \always \eventually green$, i.e.
+the light starts as red but should infinitely often be green. The system transition specification is
+%{\footnotesize
+%\begin{align}
+%\always & (\lnext ( (red \land \lnot yellow \land \lnot green) \lor (\lnot red \land yellow \land \lnot green) \lor \mbox{} \label{eq:mutex1} \\
+%	& \quad \quad \quad (\lnot red \land \lnot yellow \land green)) \land \mbox{}  \label{eq:mutex2} \\
+%	& \;\; (red \land \lnext tick \limplies \lnext green) \land (green \land \lnext tick \limplies \lnext yellow) \land \mbox{} \label{eq:change1} \\
+%	& \quad \quad \quad (yellow \land \lnext tick \limplies \lnext red) \land \mbox{}  \label{eq:change2} \\
+%	& \;\; (red \land \lnext \lnot tick \limplies red) \land (green \land \lnext \lnot tick \limplies green) \land \mbox{} \label{eq:same1} \\
+%	& \quad \quad \quad (yellow \land \lnext \lnot tick \limplies yellow) \label{eq:same2})
+%\end{align}
+%}%
+%That is, \eqnref{eq:mutex1} - \eqnref{eq:mutex2} express that the light should only be one color at a time; 
+%\eqnref{eq:change1} - \eqnref{eq:change2} that the color should change from red to green to yellow and back to red whenever $tick$ is true; 
+%and the color should remain the same whenever $tick$ is false.
+
+{\scriptsize
+\begin{align}
+\always & (\lnext ( (red \land \lnot yellow \land \lnot green) \lor (\lnot red \land yellow \land \lnot green) \lor (\lnot red \land \lnot yellow \land green)) \land \mbox{}  \label{eq:mutex} \\
+	& \; (red \land \lnext tick \limplies \lnext green) \land (green \land \lnext tick \limplies \lnext yellow) \land (yellow \land \lnext tick \limplies \lnext red) \land \mbox{}  \label{eq:change} \\
+	& \; (red \land \lnext \lnot tick \limplies red) \land (green \land \lnext \lnot tick \limplies green) \land (yellow \land \lnext \lnot tick \limplies yellow) \label{eq:same})
+\end{align}
+}
+
+\noindent That is, the light should only be one color at a time \eqnref{eq:mutex}; 
+the color should change from red $\rightarrow$ green $\rightarrow$ yellow $\rightarrow$ red $\rightarrow \ldots$ whenever $tick$ is true \eqnref{eq:change}; 
+and the color should remain the same whenever $tick$ is false \eqnref{eq:same}.
+
+\begin{figure}
+\begin{lstlisting}[language={Ada}, basicstyle=\scriptsize, numbers=left]
+package TrafficLight with SPARK_Mode is
+  type Controller is private;
+
+  type System is record
+      red: Boolean; yellow: Boolean; green: Boolean;
+    end record;
+
+  function Is_Init(C: Controller) return Boolean;
+  function Env_Init(tick: Boolean) return Boolean is (True);
+  function Sys_Init(S: System) return Boolean is 
+    (S.red and not S.yellow and not S.green) with Ghost;
+
+  function Env_Trans(C: Controller; tick: Boolean) return Boolean
+    with Pre => (not Is_Init(C));
+  function Sys_Trans(C: Controller; tick: Boolean; S: System) 
+    return Boolean with Pre => (not Is_Init(C)), Ghost;
+
+  procedure Move(C: in out Controller; tick: in Boolean; S: out System)
+    with 
+    Pre => (if Is_Init(C) then Env_Init(tick) else Env_Trans(C, tick)),
+    Contract_Cases =>
+      (Is_Init(C) => Sys_Init(S) and (not Is_Init(C)),
+       others => Sys_Trans(C'Old, tick, S) and (not Is_Init(C)));
+
+private
+  function State_To_Input_Mapping(C: Controller) return Boolean
+    with Ghost;
+  function State_To_Output_Mapping(C: Controller) return Boolean
+    with Ghost;
+
+  subtype State_Num is Integer range 1 .. 7;
+
+  type Controller is record
+      State: State_Num := State_Num'Last; tick: Boolean; S: System;
+  end record
+    with Type_Invariant => (State_To_Input_Mapping(Controller) and 
+                            State_To_Output_Mapping(Controller));
+
+end TrafficLight;
+\end{lstlisting}
+  \caption{SPARK specification for a traffic light controller.}
+  \label{fig:trafficLightSpec}
+\end{figure}
+
+\sloppy{
+In SPARK, subunits consist of a specification and a body. 
+For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
+The package specification is shown in \figref{fig:trafficLightSpec}.
+Type \lstinline{Controller} encodes a Moore machine suitable for representing the state of a synthesized controller. 
+In the public part of the specification (line 2), \lstinline{Controller} is declared as a private type so that the user cannot arbitrarily manipulate its state 
+(possibly violating the controller's GR(1) specification). 
+In the private part of the specification (lines 34-39), \lstinline{Controller} is a record that stores the controller's internal state number, 
+current input value(s), and the current output value(s). 
+Its initial state is always the last possible state number. 
+Each input and output is of type \lstinline{Boolean}.
+When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{System} (lines 4-7) or \lstinline{Environment}, respectively. 
+Here, there is only one input, so it is not wrapped in a record. 
+Since controllers are Moore machines, \lstinline{Controller} includes a type invariant (lines 38-39) that specifies that for each internal state, 
+there is exactly one set of input values that brought the controller into that state and exactly one set of output values produced in that state. 
+Functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} encode lookup tables specifying the input and output value(s) for each state, respectively.
+Both of these are declared (lines 27-30) with aspect \lstinline{Ghost}, indicating that they used for proof purposes only, 
+i.e. they will create verification conditions related to type \lstinline{Controller} but are not executable. 
+The logic for these functions is given in the body (not shown here), but for instance in state 0, \lstinline{State_To_Input_Equivalence} returns \lstinline{False}, 
+since \lstinline{tick = False} as the system enters state 0, and \lstinline{State_To_Input_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
+since $red$ is True and $yellow$ and $green$ are false as the system enters state 0.
+ }
+
+The public function \lstinline{Is_Init} (line 9) checks whether the controller is in its initial state, i.e. no inputs have yet been received. 
+The public function \lstinline{Env_Init} (line 10) checks whether input(s) satisfy $\varphi_e^i$. 
+It is implemented as an expression function, i.e. the implementation is given directly in the specification and not the body, 
+because all terms needed to define it are visible in the public part of the specification. 
+In this example, since $\varphi_e^i = \top$ as pulled from the Salty specification, this function always simply returns \lstinline{True}. 
+The public function \lstinline{Sys_Init} (lines 11-12) checks whether outputs(s) satisfy $\varphi_s^i$. 
+It is implemented as an expression function for the same reason.  
+As pulled from the Salty specification, $\varphi^s_i = red \land \lnot yellow \land \lnot green$, so this function returns the value of
+the expression \lstinline{S.red and not S.yellow and not S.green}.
+But unlike \lstinline{Is_Init} and \lstinline{Env_Init}, it is marked with aspect \lstinline{Ghost}, 
+meaning it is ``ghost code'' used only for proof and cannot be executed. 
+\lstinline{Is_Init} and \lstinline{Env_Init} are used for proof and are also executable. 
+We chose to make these executable for reasons related to the meaning of GR(1) specifications. 
+Recall that GR(1) specifications have the form $\varphi_e \limplies \varphi_s$. 
+If $\varphi_e$ ever becomes false, i.e. if the environment produces input value(s) that violate $\varphi_e$, then the specification as whole 
+is satisfied regardless of whether the system produces output value(s) that satisfy $\varphi_s$. 
+In theory, the system in this case could produce arbitrary outputs and still satisy the overall specification. 
+In practice, we believe a user would generally want to know that the environment violated its specification, 
+so that the user could either choose the system output value(s) explicitly or fall back to some other error handling routine. 
+Therefore, a user needs to be able to check inputs with \lstinline{Env_Init} if \lstinline{Is_Init} returns true, 
+which is why both are executable.
+Public functions \lstinline{Env_Trans} and \lstinline{Sys_Trans} (lines 14-17) check whether the next set of input value(s) and output value(s) 
+satisfy $\varphi_e^t$ and $\varphi_s^t$, respectively. 
+For the same reasons as above, \lstinline{Env_Trans} is executable but \lstinline{Sys_Trans} is ghost code. 
+Note that \lstinline{Env_Trans} has a precondition that the controller must not be in its initial state, 
+since $\varphi_e^t$ can depend on both the current and next set of input value(s). 
+This precondition is not necessary for \lstinline{Sys_Trans}, since it is not executable, and it will only be used to 
+assert properties in cases in which the controller is not in its initial state. 
+As with \lstinline{Env_Init} and \lstinline{Sys_Init}, the logic for these functions is pulled from the Salty specification and 
+implemented in the body (not shown), since they make use of input and output values stored in the \lstinline{Controller}, whose fields are private.
+
+\begin{figure}
+\begin{lstlisting}[language={Ada}, basicstyle=\scriptsize]  
+procedure Move(C: in out Controller; tick: in Boolean; S: out System) is
+begin
+  case C.State is
+    when 1 =>
+      case tick is
+        when False =>
+          C.State := 1; 
+          C.S.red := True; C.S.yellow := False; C.S.green := False;
+        when True =>
+          C.State := 3; 
+          C.S.red := False; C.S.yellow := False; C.S.green := True;
+        when others =>
+          raise Program_Error;
+      end case;
+       ...      
+     when 7 =>
+      case tick is
+        when False =>
+          C.State := 1; 
+          C.S.red := True; C.S.yellow := False; C.S.green := False;
+        when True => 
+          C.State := 2; 
+          C.S.red := True; C.S.yellow := False; C.S.green := False;
+        when others =>
+          raise Program_Error;
+      end case;
+  end case;
+  C.tick := tick; S := C.S;
+end Move;
+\end{lstlisting}
+  \caption{The body of the Move procedure.}
+  \label{fig:moveBody}
+\end{figure}
+
+The public procedure \lstinline{Move} (lines 19-24) transitions a controller based on its current state and 
+next set of input value(s), and it produces the next set of output value(s). 
+It has a precondition that if the controller is in its initial state, 
+inputs must satisfy $\varphi_e^i$; otherwise they must satisfy $\varphi_e^t$.
+The aspect \lstinline{Contract_Cases} specifies additional sets of preconditions paired with postconditions, 
+where the set of all preconditions must be mutually exclusive and exhaustively cover the entire input space.
+The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
+Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to variable values before evaluation of the subprogram; 
+for the right-hand side, they refer to variable values after evaluation.
+For the right-hand side, the aspect \lstinline{Old} can therefore be used to reference the value of a variable before evaluation of the subprogram. 
+Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
+the first set of output value(s) produced should satisfy $\varphi_s^i$ and the controller should no longer be in its initial state; 
+if it is not in its initial state, then the output value(s) produced should satisfy $\varphi_s^t$, 
+evaluated based on the most recent input and output values stored in C'Old, the next input value(s) just provided (in this case in \lstinline{tick}), 
+and the next output values just generated (in this case in the record \lstinline{System}). 
+This set of contract cases embodies our main proof goal, i.e. verification of system initial and transition properties from the original GR(1) specification for the controller.
+
+A fragment of the body of \lstinline{Move} is shown in \figref{fig:moveBody}. 
+Note that there are cases that can lead to \lstinline{Program_Error}. 
+This is because we programmatically use \lstinline{others} to cover all possible input combinations 
+that would not be allowed due to $\varphi_e^i$ or $\varphi_e^t$. 
+In the traffic light example, they are unnecessary because all possible combinations of input values are 
+allowed out of each state. 
+In any case, SPARK will prove that that these cases are not reachable.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+To check whether our approach has the potential to produce SPARK controllers that can be automatically verified to satisfy their original system initial and transition specifications, 
+we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, Salty, all of which make their examples publically available for download. 
+These examples -- including GR(1) specifications, sythesized SPARK packages, and SPARK analysis results -- are now all available in Salty format on the 
+Salty GitHub repository\footnote{https://github.com/GaloisInc/salty/}, including the traffic light example of the previous section.
+
+Many of the examples could be fully proven in SPARK in a matter of seconds.   
+However, many examples were surprisingly large, with 10s or even 100s of thousands of states, 
+resulting in controller bodies that are several megabytes in some cases. 
+Some of these could be analyzed but not fully proven; often, SPARK could prove 99\% of the checks required for functional correctness of the \lstinline{Move} function. 
+Others required too much memory to analyze.
+\figref{fig:timingResults} shows the amount of time needed to analyze examples as a function of approximate number of states, 
+with examples that proved to 99\% marked in red and examples that could not be analyzed set to -1 in red. 
+Results were generated on a Linux VM given 24 MB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
+
+\begin{figure}
+\caption{Timing results for example controllers in SPARK.}
+\label{fig:timingResults}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% CASE STUDIES
+% Lessons Learned
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-\label{sec:discussion}
+\section{Lessons Learned}
+\label{sec:lessonsLearned}
+
+Throughout the process of synthesizing and attempting to verify controllers in SPARK, we learned several lessons, 
+both about SPARK and about some of the finer points of GR(1) specifications.
+
+In terms of encoding SPARK controllers for verification, we originally tried to 
+[James: fill in what we tried to do in our first iteration].
+However, using case structures in the \lstinline{Move} procedure was better suited to proof because
+[James and/or Joffrey: fill in the reason(s) why case structures are better]
+
+[James and/or Laura: I want to write a little more detail about the next section]
+We also found that some examples in our database did not have any inputs, i.e. they essentially amounted to synthesizing a system independent of an environment. 
+In that case, we had specifications for a non-existent environment that were vacuous, 
+and this was causing SPARK to take an abnormally long amount of time to verify these controllers, given their relatively small size.
+
+We also originally had some examples that resulted in invalid SPARK code. 
+In particular, there were cases in the controller that were empty. 
+After some investigation, we discovered that in the mathematical representation of the controllers, there were states with no possible successors. 
+This is due to the interpretation of GR(1) specifications as a game between the environment and the system. 
+Recall that since GR(1) specifications have the form $\varphi_e \limplies \varphi_s$, if the environment violates its specification $\varphi_e$, then the system 
+can violate its specification $\varphi_s$ and still satisfy the overall specification. 
+States with no successors represent cases in which the system can ``win'' the game by transitioning to a state in which the environment is forced to violate $\varphi_e$. 
+In \sectref{sec:implementation}, we stated that we believe a user will generally want to know when the environment violates its specification. 
+Therefore, we believe that the best way to handle this would be to return some indication to the user, so they can decide what the system should do or fall back to some other error handling routine, 
+which will require some modifications to our SPARK code generator. 
+In the meantime, we were essentially able to tighten up these specifications so that there were no such states.
+[Laura and/or James: explain why this was happening in our examples.
+Also check with the VIP example: does this reveal an error in the code synthesized in other languages?
+It's possible that there are case statements that are "falling through"]
+
+[Joffrey: If there are examples that are still taking an unexplainable amount of time, it'd be nice to have an intuition for why.]
+
+[Joffrey: It would also be nice if there were an intuition for why SPARK is generally able to exactly 99\% of checks for large examples.
+Is it the "final" form of the post condition that is causing problems?]
 
@@
 %
+\newpage
 \bibliographystyle{splncs04}

### 2020-01-21 src/CodeGen/SPARK.hs
Fix invariant violation for specs with empty environments

@@ -186,7 +186,8 @@ sparkFSM FSM { fsmName, fsmEnums, fsmInputs, fsmOutputs, fsmInitial, fsmNodes } 
                   , subBody (text "Move")
                       [ ]
                       [ mkNested fsmNodes cntlrName stName env sys
-                      , statement $ assign (qualify (Just cntlrName) (svName env)) (svName env)
+                      , statement $ assign (qualify (Just cntlrName) (svName env))
+                          (if Map.null fsmInputs then sparkValue (VBool True) else svName env)
                       , statement $ assign (svName sys) (qualify (Just cntlrName) (svName sys)) ] ] ]
 
 -- NOTE: Ada is case-insensitive

### 2020-01-22 doc/2020-CAV/paper.tex
Draft paper updates

@@
 LinQuest Corp., Beavercreek, OH 45431, USA \\
-\email{james.hamil@ctr.us.af.mil}  \and
+\email{james.hamil.ctr@us.af.mil}  \and
 AdaCore, F-75009 Paris, France \\
@@
 a subset of properties comprising their original GR(1) specifications, namely system initial and transition properties. 
-We also discuss leasons learned about how to best encode controllers in SPARK for verification, 
+We also discuss lessons learned about how to best encode controllers in SPARK for verification, 
 examples in which verification found unexpected controller behaviors, and 
@@
 Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
-a back-end that turns controller designs into executable sofware implementations in a variety of programming languages.
+a back-end that turns controller designs into executable software implementations in a variety of programming languages.
 
@@
 synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their functional specifications in SPARK. 
-To date, we have largely achieved this goal for a subset of properties comprising GR(1) specifiations, 
+To date, we have largely achieved this goal for a subset of properties comprising GR(1) specifications, 
 i.e. system initial and transition properties, for moderately-sized controllers.
@@
 i.e. they will create verification conditions related to type \lstinline{Controller} but are not executable. 
-The logic for these functions is given in the body (not shown here), but for instance in state 0, \lstinline{State_To_Input_Equivalence} returns \lstinline{False}, 
-since \lstinline{tick = False} as the system enters state 0, and \lstinline{State_To_Input_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
-since $red$ is True and $yellow$ and $green$ are false as the system enters state 0.
+The logic for these functions is given in the body (not shown here), but for instance in state 1, \lstinline{State_To_Input_Equivalence} returns \lstinline{False}, 
+since \lstinline{tick = False} as the system enters state 1, and \lstinline{State_To_Input_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
+since $red$ is True and $yellow$ and $green$ are false as the system enters state 1.
  }
@@
 is satisfied regardless of whether the system produces output value(s) that satisfy $\varphi_s$. 
-In theory, the system in this case could produce arbitrary outputs and still satisy the overall specification. 
+In theory, the system in this case could produce arbitrary outputs and still satisfy the overall specification. 
 In practice, we believe a user would generally want to know that the environment violated its specification, 
@@
 that would not be allowed due to $\varphi_e^i$ or $\varphi_e^t$. 
+This makes the code generation simpler to implement as each variable's domain does not need to be explicitly tracked and covered. 
 In the traffic light example, they are unnecessary because all possible combinations of input values are 
 allowed out of each state. 
-In any case, SPARK will prove that that these cases are not reachable.
+In any case, SPARK will prove that these cases are not reachable.
 
@@
 To check whether our approach has the potential to produce SPARK controllers that can be automatically verified to satisfy their original system initial and transition specifications, 
-we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, Salty, all of which make their examples publically available for download. 
-These examples -- including GR(1) specifications, sythesized SPARK packages, and SPARK analysis results -- are now all available in Salty format on the 
+we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, Salty, all of which make their examples publicly available for download. 
+These examples -- including GR(1) specifications, synthesized SPARK packages, and SPARK analysis results -- are now all available in Salty format on the 
 Salty GitHub repository\footnote{https://github.com/GaloisInc/salty/}, including the traffic light example of the previous section.
@@
 
+[Laura: Summarize paper.
+Highlight benefits of SPARK, especially if it revealed "fall through" case statements in some examples.
+Mention it was relatively easy to figure out how to structure controllers
+and annotate code]
+
+[Joffrey: Speculate on how to prove liveness, whether in SPARK and/or using other tools/approaches.]
+
 %

### 2020-01-25 doc/2020-CAV/paper.tex
Paper edits to incorporate Joffrey's edits, James's edits on Moore machines, definitions of temporal operators

@@
 
-\noindent It is assumed that at each time step, the environment chooses an input from $\inputs$, then the system chooses an output from $\outputs$ in response. 
+\noindent 
+In this context, the temporal operators $\always$ ``always,'' $\eventually$ ``eventually,'' and $\lnext$  ``next'' have the following meanings. 
+A formula of the form $\always b$ holds if $b$ is true at every time step. 
+A formula of the form $\eventually b$ holds if $b$ is eventually true at some future time step. 
+A formula of the form $\always \eventually b$ holds if $b$ is true infinitely often in the future. 
+And a formula of the form $\lnext b$ holds if $b$ is true at the next time step. 
+It is assumed that at each time step, the environment chooses an input from $\inputs$, then the system chooses an output from $\outputs$ in response. 
 The result of synthesis is a \emph{control protocol} or \emph{strategy} that can be expressed as a Moore machine, 
@@
 Here, there is only one input, so it is not wrapped in a record. 
-Since controllers are Moore machines, \lstinline{Controller} includes a type invariant (lines 38-39) that specifies that for each internal state, 
-there is exactly one set of input values that brought the controller into that state and exactly one set of output values produced in that state. 
-Functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} encode lookup tables specifying the input and output value(s) for each state, respectively.
-Both of these are declared (lines 27-30) with aspect \lstinline{Ghost}, indicating that they used for proof purposes only, 
-i.e. they will create verification conditions related to type \lstinline{Controller} but are not executable. 
+Since \lstinline{Controller} is a Moore machine, for each internal state, 
+there is exactly one set of output values produced in that state. 
+Furthermore, controllers synthesized from GR(1) specifications have the additional property that 
+there is exactly one set of input values that brought the controller into that state. 
+Therefore, \lstinline{Controller} includes a type invariant (lines 38-39) that captures this property, 
+where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
+encode lookup tables specifying the input and output value(s) for each state, respectively.
+Both of these are declared (lines 27-30) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
+i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
+unless it is specified with a certain switch during compilation that they should be.
+Executing ghost code is often used for debugging, e.g. to check whether an unproved property is actually true. 
+In this case, removing the ghost code can save a lot of memory, 
+especially on bigger examples where the tables encoded by functions  \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
+would have to store information on thousands of states.
 The logic for these functions is given in the body (not shown here), but for instance in state 1, \lstinline{State_To_Input_Equivalence} returns \lstinline{False}, 
-since \lstinline{tick = False} as the system enters state 1, and \lstinline{State_To_Input_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
+since \lstinline{tick = False} as the system enters state 1, and \lstinline{State_To_Output_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
 since $red$ is True and $yellow$ and $green$ are false as the system enters state 1.
@@
 But unlike \lstinline{Is_Init} and \lstinline{Env_Init}, it is marked with aspect \lstinline{Ghost}, 
-meaning it is ``ghost code'' used only for proof and cannot be executed. 
-\lstinline{Is_Init} and \lstinline{Env_Init} are used for proof and are also executable. 
-We chose to make these executable for reasons related to the meaning of GR(1) specifications. 
+meaning it is ``ghost code'' used only for proof and will not be executed.
+\lstinline{Is_Init} and \lstinline{Env_Init} are used for proof but are also callable in functional code. 
+We chose to make them non-ghost functions for reasons related to the meaning of GR(1) specifications. 
 Recall that GR(1) specifications have the form $\varphi_e \limplies \varphi_s$. 
@@
 Therefore, a user needs to be able to check inputs with \lstinline{Env_Init} if \lstinline{Is_Init} returns true, 
-which is why both are executable.
+which is why both are callable. At the moment, they are not used in the code of the \lstinline{Move} procedure.
 Public functions \lstinline{Env_Trans} and \lstinline{Sys_Trans} (lines 14-17) check whether the next set of input value(s) and output value(s) 
 satisfy $\varphi_e^t$ and $\varphi_s^t$, respectively. 
-For the same reasons as above, \lstinline{Env_Trans} is executable but \lstinline{Sys_Trans} is ghost code. 
+For the same reasons as above, \lstinline{Sys_Trans} is a ghost function but \lstinline{Env_Trans} is not.
 Note that \lstinline{Env_Trans} has a precondition that the controller must not be in its initial state, 
 since $\varphi_e^t$ can depend on both the current and next set of input value(s). 
-This precondition is not necessary for \lstinline{Sys_Trans}, since it is not executable, and it will only be used to 
-assert properties in cases in which the controller is not in its initial state. 
+This precondition is also necessary for \lstinline{Sys_Trans}, because in the initial state, the attributes of the \lstinline{Controller} are not initialized, and the \lstinline{Sys_Trans} function might read these values.
 As with \lstinline{Env_Init} and \lstinline{Sys_Init}, the logic for these functions is pulled from the Salty specification and 

### 2020-01-28 doc/2020-CAV/paper.tex
Edits to get to final draft

@@
 %
-\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
-James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
-Joffrey Heugenot\inst{3}\orcidID{2222--3333-4444-5555}}
+%\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
+%James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
+%Joffrey Huguet\inst{3}\orcidID{2222--3333-4444-5555}}
+\author{Laura Humphrey\inst{1} \and
+James Hamil\inst{2} \and
+Joffrey Huguet\inst{3}}
 %
@@
 from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
-SPARK is both a language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.  
+SPARK is both a programming language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.
 In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
-a subset of properties comprising their original GR(1) specifications, namely system initial and transition properties. 
+a subset of properties comprising GR(1) specifications, namely system initial and transition properties. 
 We also discuss lessons learned about how to best encode controllers in SPARK for verification, 
@@
  $\varphi^e_t$, $\varphi^s_t$ & - &  Formulas of the form $\bigwedge_{j \in J} \always B_j$, where each $B_j$ is a Boolean combination of variables from $\inputs \cup \outputs$ and expressions of the form $\lnext v$, where $v \in \inputs$ for $\varphi^e_t$ and $v \in \inputs \cup \outputs$ for $\varphi^s_t$.
-These encode properties that should always hold as well as rules for how inputs and outputs are allowed to change based on current input and output values.
+These encode properties that should always hold as well as rules for how inputs and outputs are allowed to change based on most recent input and output values.
 \end{xtabular}
@@
 In this context, the temporal operators $\always$ ``always,'' $\eventually$ ``eventually,'' and $\lnext$  ``next'' have the following meanings. 
-A formula of the form $\always b$ holds if $b$ is true at every time step. 
-A formula of the form $\eventually b$ holds if $b$ is eventually true at some future time step. 
-A formula of the form $\always \eventually b$ holds if $b$ is true infinitely often in the future. 
-And a formula of the form $\lnext b$ holds if $b$ is true at the next time step. 
+A formula of the form $\always b$ holds if $b$ is true at every time step, 
+$\eventually b$ holds if $b$ is eventually true at some future time step, 
+$\always \eventually b$ holds if $b$ is true infinitely often in the future, 
+and $\lnext b$ holds if $b$ is true at the next time step. 
 It is assumed that at each time step, the environment chooses an input from $\inputs$, then the system chooses an output from $\outputs$ in response. 
-The result of synthesis is a \emph{control protocol} or \emph{strategy} that can be expressed as a Moore machine, 
-with each state encoding the most recent input values from the environment and the most recent output values produced in response by the system.
+The result of synthesis is a \emph{control protocol} or \emph{strategy} that can be expressed as a Moore machine. 
+In a Moore machine, each state produces a unique set of system output values, 
+and in the case of GR(1), synthesized Moore machines have the additional property that 
+there is a unique set of environment input values that brings the system into each state.
 
@@
 synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their functional specifications in SPARK. 
-To date, we have largely achieved this goal for a subset of properties comprising GR(1) specifications, 
+To date, we have achieved this goal for a subset of properties comprising GR(1) specifications, 
 i.e. system initial and transition properties, for moderately-sized controllers.
 In what follows, in \sectref{sec:implementation} we discuss the SPARK implementation. 
-In \sectref{sec:caseStudies}, we demonstrate the efficacy of our approach on examples pulled from a variety of sources.
-In \sectref{sec:discussion}, we discuss possible approaches for proving liveness properties as well as design choice alternatives, 
-e.g. for dealing with environment specifications.
-We end with concluding remarks in \sectref{sec:conclusions}.
+In \sectref{sec:caseStudies}, we demonstrate the efficacy of our approach on examples pulled from various sources.
+In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to structure synthesized controllers for proof, 
+examples in which SPARK revealed unexpected controller behaviors, and caveats related to the interpretation of GR(1) specifications. 
+We end with concluding remarks in \sectref{sec:conclusions}, including a discussion of future work.
 
@@
 Let this input be called $tick$.
-The traffic light's color changes from red to green to yellow and back to red, repeating the cycle infinitely. 
+The traffic light's color changes from red to green to yellow to red $\ldots$, repeating the cycle infinitely. 
 Let the output variables then be $red$, $yellow$, and $green$.
@@
 For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
-The package specification is shown in \figref{fig:trafficLightSpec}.
+The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
 Type \lstinline{Controller} encodes a Moore machine suitable for representing the state of a synthesized controller. 
-In the public part of the specification (line 2), \lstinline{Controller} is declared as a private type so that the user cannot arbitrarily manipulate its state 
+In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state 
 (possibly violating the controller's GR(1) specification). 
-In the private part of the specification (lines 34-39), \lstinline{Controller} is a record that stores the controller's internal state number, 
+In the private part of the specification, \lstinline{Controller} (lines 34-39) is a record that stores the controller's internal state number, 
 current input value(s), and the current output value(s). 
@@
 Each input and output is of type \lstinline{Boolean}.
-When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{System} (lines 4-7) or \lstinline{Environment}, respectively. 
+When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{System} (lines 4-6) or \lstinline{Environment}, respectively. 
 Here, there is only one input, so it is not wrapped in a record. 
@@
 encode lookup tables specifying the input and output value(s) for each state, respectively.
-Both of these are declared (lines 27-30) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
+Both of these are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
 i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
@@
 Executing ghost code is often used for debugging, e.g. to check whether an unproved property is actually true. 
-In this case, removing the ghost code can save a lot of memory, 
+In this case, removing the ghost code can save significant memory, 
 especially on bigger examples where the tables encoded by functions  \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
@@
 
-The public function \lstinline{Is_Init} (line 9) checks whether the controller is in its initial state, i.e. no inputs have yet been received. 
-The public function \lstinline{Env_Init} (line 10) checks whether input(s) satisfy $\varphi_e^i$. 
+The public function \lstinline{Is_Init} (line 8) checks whether the controller is in its initial state, i.e. no inputs have yet been received. 
+The public function \lstinline{Env_Init} (line 9) checks whether input(s) satisfy $\varphi_e^i$. 
 It is implemented as an expression function, i.e. the implementation is given directly in the specification and not the body, 
@@
 In this example, since $\varphi_e^i = \top$ as pulled from the Salty specification, this function always simply returns \lstinline{True}. 
-The public function \lstinline{Sys_Init} (lines 11-12) checks whether outputs(s) satisfy $\varphi_s^i$. 
+The public function \lstinline{Sys_Init} (lines 10-11) checks whether outputs(s) satisfy $\varphi_s^i$. 
 It is implemented as an expression function for the same reason.  
@@
 the expression \lstinline{S.red and not S.yellow and not S.green}.
-But unlike \lstinline{Is_Init} and \lstinline{Env_Init}, it is marked with aspect \lstinline{Ghost}, 
-meaning it is ``ghost code'' used only for proof and will not be executed.
+But unlike \lstinline{Is_Init} and \lstinline{Env_Init}, it is marked with aspect \lstinline{Ghost} because it is mainly used for proof and does not need to be executed. 
 \lstinline{Is_Init} and \lstinline{Env_Init} are used for proof but are also callable in functional code. 
-We chose to make them non-ghost functions for reasons related to the meaning of GR(1) specifications. 
+We chose to make these non-ghost functions for reasons related to the meaning of GR(1) specifications. 
 Recall that GR(1) specifications have the form $\varphi_e \limplies \varphi_s$. 
@@
 which is why both are callable. At the moment, they are not used in the code of the \lstinline{Move} procedure.
-Public functions \lstinline{Env_Trans} and \lstinline{Sys_Trans} (lines 14-17) check whether the next set of input value(s) and output value(s) 
+Public functions \lstinline{Env_Trans} and \lstinline{Sys_Trans} (lines 13-16) check whether the next set of input value(s) and output value(s) 
 satisfy $\varphi_e^t$ and $\varphi_s^t$, respectively. 
@@
 
-The public procedure \lstinline{Move} (lines 19-24) transitions a controller based on its current state and 
+The public procedure \lstinline{Move} (lines 18-23) transitions a controller based on its current state and 
 next set of input value(s), and it produces the next set of output value(s). 
@@
 if it is not in its initial state, then the output value(s) produced should satisfy $\varphi_s^t$, 
-evaluated based on the most recent input and output values stored in C'Old, the next input value(s) just provided (in this case in \lstinline{tick}), 
+evaluated based on the most recent input and output values stored in \lstinline{C'Old}, the next input value(s) just provided (in this case in \lstinline{tick}), 
 and the next output values just generated (in this case in the record \lstinline{System}). 
@@
 Note that there are cases that can lead to \lstinline{Program_Error}. 
-This is because we programmatically use \lstinline{others} to cover all possible input combinations 
+This is because case statements require all possible cases to be covered, so 
+we programmatically use \lstinline{others} to cover all possible input combinations 
 that would not be allowed due to $\varphi_e^i$ or $\varphi_e^t$. 
-This makes the code generation simpler to implement as each variable's domain does not need to be explicitly tracked and covered. 
 In the traffic light example, they are unnecessary because all possible combinations of input values are 
@@
 In any case, SPARK will prove that these cases are not reachable.
+
+We briefly note that Salty includes language features that can result in different constructs being used 
+to represent inputs and outputs in synthesized controllers, 
+including enumerations and integers. 
+For instance, enumerations encode that an enumerated input or output can have exactly one of a set of values, 
+as in the traffic light being exactly one color as expressed in $\varphi_s^i$ and part \eqnref{eq:mutex} of $\varphi_s^t$. 
+In such cases, enumerations or integers can make both specifications and code more compact and easier to read and understand.
 
@@
 To check whether our approach has the potential to produce SPARK controllers that can be automatically verified to satisfy their original system initial and transition specifications, 
-we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, Salty, all of which make their examples publicly available for download. 
+we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, and Salty, all of which make their examples publicly available for download. 
 These examples -- including GR(1) specifications, synthesized SPARK packages, and SPARK analysis results -- are now all available in Salty format on the 
@@
 
-Many of the examples could be fully proven in SPARK in a matter of seconds.   
-However, many examples were surprisingly large, with 10s or even 100s of thousands of states, 
-resulting in controller bodies that are several megabytes in some cases. 
-Some of these could be analyzed but not fully proven; often, SPARK could prove 99\% of the checks required for functional correctness of the \lstinline{Move} function. 
-Others required too much memory to analyze.
 \figref{fig:timingResults} shows the amount of time needed to analyze examples as a function of approximate number of states, 
-with examples that proved to 99\% marked in red and examples that could not be analyzed set to -1 in red. 
+with examples with unproved functional contracts marked in red and examples that could not be analyzed set to -10 in red.
 Results were generated on a Linux VM given 24 MB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
+On most examples in which the Moore machine encoding the controller had less than 4000 transitions, 
+SPARK was able to completely verify or prove the generated code complies with its specification. 
+Examples that had more than 4000 transitions required too much memory to analyze, resulting in errors when attempting to verify them in SPARK. 
+Of examples with less than 4000 transitions, two resulted in errors. 
+These two examples had abnormally large specifications consisting of approximately 1000 terms each, whereas most other examples had 500 or less.
+Two could only be partially proven. 
+One was relatively large, with about 2000 transitions. 
+The other had arithmetic terms in its specification (Salty and Slugs support integer inputs and outputs). 
+We are investigating why this example does not fully prove, 
+since we feel that SPARK should be capable of fully proving this example.
 
 \begin{figure}
-\caption{Timing results for example controllers in SPARK.}
+\includegraphics[width=\textwidth]{salty.eps}
+\caption{Timing results for example controllers in SPARK as a function of number of transitions in the Moore machine representing the controller.
+Examples marked ``proven'' were fully verified, examples marked ``partial'' were partially verified, and examples marked ``error'' were too big to analyze.}
 \label{fig:timingResults}
@@
 In terms of encoding SPARK controllers for verification, we originally tried to 
-[James: fill in what we tried to do in our first iteration].
-However, using case structures in the \lstinline{Move} procedure was better suited to proof because
-[James and/or Joffrey: fill in the reason(s) why case structures are better]
+mirror the approach taken in other Salty language targets by building a static lookup table for state transitions. 
+To do this, we tried to create an array of \lstinline{Formal_Hashed_Maps} indexed by \lstinline{State_Num}, where keys 
+were derived from environment input values and used to look up the next \lstinline{State_Num} and corresponding system output values. 
+Ghost functions consisting of nested quantified expressions were used to assert that in each state, initial and transition properties held 
+according to values stored in the current state's hashed map and the outputs of next states. 
+These functions comprised the post-condition of a function that initialized the controller's lookup table. 
+The body of the \lstinline{Move} procedure simply retrieved the outputs and next state from the lookup table 
+using its stored \lstinline{State_Num} and \lstinline{Environment} input.
+The public portion of the SPARK specification was largely unchanged.
+This approach was only able to prove the smallest of examples in a reasonable amount of time.
 
-[James and/or Laura: I want to write a little more detail about the next section]
-We also found that some examples in our database did not have any inputs, i.e. they essentially amounted to synthesizing a system independent of an environment. 
-In that case, we had specifications for a non-existent environment that were vacuous, 
+While the use of formal containers was intuitive, they are more complex to reason about with respect to proof 
+because they require reasoning about models of the containers and the effects of container operations. 
+Encoding the the lookup table as a case statement is more straightforward, because for instance, 
+it is ``obvious'' to the underlying solvers that state transitions are static and that a transition exists for every possible input, since 
+case statements must be exhaustive. 
+%When building a lookup table represented as a formal container, one can imagine the provers would have to do additional reasoning 
+%to ensure that all possible transitions are defined for each state, and no entries are ever deleted or added to the container at other points in the code. 
+Encoding the \lstinline{Move} procedure as a case statement still has some issues, mainly that 
+(1) the generated code can be quite long, leading to memory errors when trying to prove the subprogram with SPARK, and
+(2) when proving the postcondition, since the solvers prove all cases at the same time and the number of cases grows 
+exponentially with the number of inputs, sometimes the solvers are not able to prove the postcondition.
+A solution to both problems could be to split the \lstinline{Move} procedure into several smaller procedures. %, e.g. splitting on the number of states 
+% in a manner similar to a classic programmer splitting large subprograms into smaller ones. 
+This would allow SPARK to apply its modular analysis on several smaller procedures, thus enabling the proof on larger files. 
+We are currently investigating ways to split up the procedure that does not accidentally create more difficulties for the solvers.
+
+Analysis with SPARK did reveal unexpected behaviors in two controllers synthesized from our corpus of examples. 
+In particular, these controllers contained empty case statements in the \lstinline{Move} procedure. 
+Upon inspection of the Slugs output, these corresponded to states with no successors. 
+Upon inspection of the specifications, we found that these two examples\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} had 
+environment transition terms that allowed the environment to produce an input that would always cause 
+the environment to violate its own specification on the next time step. 
+These specifications had the form $\always \lnot p$ instead of $\always \lnext \lnot p$. 
+The distinction is important. 
+In the former case, the environment can choose $p$ for its next input and not violate the specification on the current time step. 
+However, it violates it on the next time step when $p$ becomes the ``current'' value instead of the ``next'' value. 
+The latter case does not allow $p$ on the next time step, and so it also cannot hold on future ``current'' steps. 
+Generally, specifications should follow the latter form, 
+and this was indeed an error in these specifications, so we fixed them.
+However, we also modified Salty to put \lstinline{raise Program_Error} for such cases and checked 
+that we were still able to prove $\varphi_e \limplies \varphi_s$ for these examples  
+(since reaching these cases would require violating the precondition $\varphi_e$ on the environment). 
+We found that in other Salty language targets, procedures equivalent to \lstinline{Move} would have raised random runtime exceptions 
+in such cases due to empty entries in their lookup tables.
+
+Based on our first round of SPARK analysis, we also found some examples in our database that did not have any inputs, 
+i.e. they essentially amounted to synthesizing a system independent of an environment. 
+In those cases, we had specifications for a non-existent environment that were vacuous, 
 and this was causing SPARK to take an abnormally long amount of time to verify these controllers, given their relatively small size. 
-
-We also originally had some examples that resulted in invalid SPARK code. 
-In particular, there were cases in the controller that were empty. 
-After some investigation, we discovered that in the mathematical representation of the controllers, there were states with no possible successors. 
-This is due to the interpretation of GR(1) specifications as a game between the environment and the system. 
-Recall that since GR(1) specifications have the form $\varphi_e \limplies \varphi_s$, if the environment violates its specification $\varphi_e$, then the system 
-can violate its specification $\varphi_s$ and still satisfy the overall specification. 
-States with no successors represent cases in which the system can ``win'' the game by transitioning to a state in which the environment is forced to violate $\varphi_e$. 
-In \sectref{sec:implementation}, we stated that we believe a user will generally want to know when the environment violates its specification. 
-Therefore, we believe that the best way to handle this would be to return some indication to the user, so they can decide what the system should do or fall back to some other error handling routine, 
-which will require some modifications to our SPARK code generator. 
-In the meantime, we were essentially able to tighten up these specifications so that there were no such states.
-[Laura and/or James: explain why this was happening in our examples.
-Also check with the VIP example: does this reveal an error in the code synthesized in other languages?
-It's possible that there are case statements that are "falling through"]
-
-[Joffrey: If there are examples that are still taking an unexplainable amount of time, it'd be nice to have an intuition for why.]
-
-[Joffrey: It would also be nice if there were an intuition for why SPARK is generally able to exactly 99\% of checks for large examples.
-Is it the "final" form of the post condition that is causing problems?]
+These controllers were did not have errors per se, 
+but they were inefficiently encoded. 
+We modified Salty to handle such cases by removing the environment, functions over the environment, and
+all references to the environment in all pre- and postconditions.
+This greatly decreased verification time and also reduced the size and increased the efficiency of the code.
 
@@
 
-[Laura: Summarize paper.
-Highlight benefits of SPARK, especially if it revealed "fall through" case statements in some examples.
-Mention it was relatively easy to figure out how to structure controllers
-and annotate code]
+We were able to successfully use SPARK to verify safety and transition properties of moderately sized controllers 
+synthesized from GR(1) specifications by the tool Salty. 
+Encoding the controllers and all of the annotations necessary for these controllers to prove automatically was relatively straighforward. 
+The act of performing ``end-to-end'' verification with SPARK on such controllers was valuable because (1) it revealed implementation errors for certain 
+types of specifications that result in random runtime errors in other Salty target languages, and (2) it revealed cases in which controllers were inefficiently encoded, i.e.
+when there is no environment.
 
-[Joffrey: Speculate on how to prove liveness, whether in SPARK and/or using other tools/approaches.]
+In terms of future work, we can potentially improve the scalability of our approach by decomposing the \lstinline{Move} procedure 
+into several subprocedures, as discussed in the previous section. 
+We are also interested in expressing and proving liveness properties. 
+Liveness properties will be more challenging to verify because they necessarily require reasoning about future states. 
+Verifying system liveness in SPARK will require something like encoding a lookahead and showing that certain states will inevitably be reached 
+when the environment satisfies its specification, which can also include liveness terms. 
+This is likely to result in complex first-order formulas with alternating quantification over time, which are 
+notoriously hard to handle for automated solvers, so discharging the resulting proof obligations may prove to be a challenge.
+To tackle this issue, collaboration with a model checker performing verification at the level of the input language might be appropriate.
 

### 2020-01-28 doc/2020-CAV/paper.tex
Minor paper edits mostly by Joffrey

@@
 (possibly violating the controller's GR(1) specification). 
-In the private part of the specification, \lstinline{Controller} (lines 34-39) is a record that stores the controller's internal state number, 
+In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
 current input value(s), and the current output value(s). 
@@
 Each input and output is of type \lstinline{Boolean}.
-When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{System} (lines 4-6) or \lstinline{Environment}, respectively. 
+When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
 Here, there is only one input, so it is not wrapped in a record. 
@@
 there is exactly one set of input values that brought the controller into that state. 
-Therefore, \lstinline{Controller} includes a type invariant (lines 38-39) that captures this property, 
-where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
-encode lookup tables specifying the input and output value(s) for each state, respectively.
-Both of these are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
+Therefore, \lstinline{Controller} includes a type invariant (lines 36-37) that captures this property, 
+where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} state whether the controller has the corresponding state number and input or output value(s), respectively.
+They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
+Both of the functions used as invariants are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
 i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
@@
 In this case, removing the ghost code can save significant memory, 
-especially on bigger examples where the tables encoded by functions  \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
+especially on bigger examples where the tables encoded for functions  \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
 would have to store information on thousands of states.
@@
 Recall that GR(1) specifications have the form $\varphi_e \limplies \varphi_s$. 
-If $\varphi_e$ ever becomes false, i.e. if the environment produces input value(s) that violate $\varphi_e$, then the specification as whole 
+If $\varphi_e$ ever becomes false, i.e. if the environment produces input value(s) that violate $\varphi_e$, then the specification as a whole 
 is satisfied regardless of whether the system produces output value(s) that satisfy $\varphi_s$. 
@@
 
-\figref{fig:timingResults} shows the amount of time needed to analyze examples as a function of approximate number of states, 
-with examples with unproved functional contracts marked in red and examples that could not be analyzed set to -10 in red.
-Results were generated on a Linux VM given 24 MB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
+\figref{fig:timingResults} shows the amount of time needed to analyze examples as a function of number of transitions in the Moore machine representing the controller, with examples that could not be analyzed due to memory errors set to -10.
+Results were generated on a Linux VM given 24 GB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
 On most examples in which the Moore machine encoding the controller had less than 4000 transitions, 
@@
 One was relatively large, with about 2000 transitions. 
-The other had arithmetic terms in its specification (Salty and Slugs support integer inputs and outputs). 
-We are investigating why this example does not fully prove, 
+The other had arithmetic terms in its specification (Salty and Slugs support integer inputs and outputs); 
+we are investigating why this example does not fully prove, 
 since we feel that SPARK should be capable of fully proving this example.
@@
 were derived from environment input values and used to look up the next \lstinline{State_Num} and corresponding system output values. 
-Ghost functions consisting of nested quantified expressions were used to assert that in each state, initial and transition properties held 
+Ghost functions consisting of nested quantified expressions were used to check that in each state, initial and transition properties held 
 according to values stored in the current state's hashed map and the outputs of next states. 
@@
 because they require reasoning about models of the containers and the effects of container operations. 
-Encoding the the lookup table as a case statement is more straightforward, because for instance, 
+Encoding the lookup table as a case statement is more straightforward, because for instance, 
 it is ``obvious'' to the underlying solvers that state transitions are static and that a transition exists for every possible input, since 

### 2020-01-28 doc/2020-CAV/paper.tex
Edits to free up space

@@
 The result of synthesis is a \emph{control protocol} or \emph{strategy} that can be expressed as a Moore machine. 
-In a Moore machine, each state produces a unique set of system output values, 
-and in the case of GR(1), synthesized Moore machines have the additional property that 
-there is a unique set of environment input values that brings the system into each state.
+%Note that in the case of GR(1), synthesized Moore machines not only produce a single set of system output values in each state, 
+%there is a unique set of environment input values that brings the controller into each state.
 
@@
 For example, RATSY %(Requirements Analysis Tool with Synthesis) 
-\cite{bloem2010ratsy} has a focus on circuit design and can synthesize designs 
+\cite{bloem2010ratsy} has a focus on circuit design and synthesizes designs 
 encoded in BLIF, %(Berkeley Logic Interchange Format),
@@
 %(SmalL bUt Complete GROne Synthesizer) 
-\cite{Ehlers2016} is architected to allow users to easily tailor synthesis algorithms for specific applications,
+\cite{Ehlers2016} is architected to allow users to tailor synthesis algorithms,
  e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
@@
 tool implementation errors could still result in errors in synthesized controllers. 
-We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications. 
+We are therefore interested in ``end-to-end'' verification of synthesized controllers. % with respect to their original GR(1) specifications. 
 Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK. 
@@
 However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
-synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their functional specifications in SPARK. 
+synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their GR(1) specifications in SPARK. 
 To date, we have achieved this goal for a subset of properties comprising GR(1) specifications, 
@@
 In what follows, in \sectref{sec:implementation} we discuss the SPARK implementation. 
-In \sectref{sec:caseStudies}, we demonstrate the efficacy of our approach on examples pulled from various sources.
+In \sectref{sec:caseStudies}, we discuss our results on examples pulled from various sources.
 In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to structure synthesized controllers for proof, 
-examples in which SPARK revealed unexpected controller behaviors, and caveats related to the interpretation of GR(1) specifications. 
+examples in which SPARK revealed unexpected controller behaviors, and caveats related to GR(1) specifications. 
 We end with concluding remarks in \sectref{sec:conclusions}, including a discussion of future work.
@@
 
-To explain the structure of the controllers we generate in SPARK, consider a simple example of a controller for a traffic light. 
-Color changes are triggered in every state in which an input signal's value is true.
-Let this input be called $tick$.
-The traffic light's color changes from red to green to yellow to red $\ldots$, repeating the cycle infinitely. 
+To explain the structure of our synthesized SPARK controllers, consider a simple example of a traffic light controller. 
+Let the controller's single input be $tick$. 
+Color changes occur in every state in which $tick$ is true.
+The traffic light's color cycles infinitely over the sequence red $\rightarrow$ green $\rightarrow$ yellow $\rightarrow$ red $\rightarrow$ $\ldots$.
 Let the output variables then be $red$, $yellow$, and $green$.
@@
 and the color should remain the same whenever $tick$ is false \eqnref{eq:same}.
+
+\renewcommand{\floatpagefraction}{.8}%
 
@@
 
-\sloppy{
-In SPARK, subunits consist of a specification and a body. 
-For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
+In SPARK, subunits consist of a specification and a body, 
+which Salty generates. 
+% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
 The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
-Type \lstinline{Controller} encodes a Moore machine suitable for representing the state of a synthesized controller. 
-In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state 
-(possibly violating the controller's GR(1) specification). 
+Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
+In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
+% (possibly violating the controller's GR(1) specification). 
 In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
@@
 where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} state whether the controller has the corresponding state number and input or output value(s), respectively. 
-They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
-Both of the functions used as invariants are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
+% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
+Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
 i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
@@
 In this case, removing the ghost code can save significant memory, 
-especially on bigger examples where the tables encoded for functions  \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
-would have to store information on thousands of states.
-The logic for these functions is given in the body (not shown here), but for instance in state 1, \lstinline{State_To_Input_Equivalence} returns \lstinline{False}, 
-since \lstinline{tick = False} as the system enters state 1, and \lstinline{State_To_Output_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
-since $red$ is True and $yellow$ and $green$ are false as the system enters state 1.
- }
+since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
+internally encode large lookup tables that have to store information on every state of the controller, 
+which is often thousands of states or even millions of states for the largest example in our database.
+The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
+\lstinline{State_To_Input_Mapping} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
+\lstinline{State_To_Input_Mapping} returns \lstinline{True} if and only if 
+\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
+since \lstinline{tick = False} brings the system into state 1 and $red$ is true and $yellow$ and $green$ are false in state 1.
 
@@
 The public function \lstinline{Env_Init} (line 9) checks whether input(s) satisfy $\varphi_e^i$. 
-It is implemented as an expression function, i.e. the implementation is given directly in the specification and not the body, 
+It is implemented as an expression function, i.e. the implementation is given directly in the specification, 
 because all terms needed to define it are visible in the public part of the specification. 
@@
 The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
-Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to variable values before evaluation of the subprogram; 
-for the right-hand side, they refer to variable values after evaluation.
-For the right-hand side, the aspect \lstinline{Old} can therefore be used to reference the value of a variable before evaluation of the subprogram. 
+Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
+for the right-hand side, they refer to values after evaluation.
+Therefore, the aspect \lstinline{Old} can be used on the right to reference the value of a variable before evaluation of the subprogram. 
 Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
@@
 that would not be allowed due to $\varphi_e^i$ or $\varphi_e^t$. 
-In the traffic light example, they are unnecessary because all possible combinations of input values are 
+In the traffic light example, these are unnecessary because all possible combinations of input values are 
 allowed out of each state. 
-In any case, SPARK will prove that these cases are not reachable.
+In any case, SPARK will prove that such cases are not reachable.
 

### 2020-01-28 doc/2020-CAV/paper.tex
Laura edits

@@
 we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, and Salty, all of which make their examples publicly available for download. 
-These examples -- including GR(1) specifications, synthesized SPARK packages, and SPARK analysis results -- are now all available in Salty format on the 
+These examples -- including GR(1) specifications in Salty format, synthesized SPARK packages, and SPARK analysis results -- are all available on the 
 Salty GitHub repository\footnote{https://github.com/GaloisInc/salty/}, including the traffic light example of the previous section. 
+We note that while some examples are small and simple, e.g. demos along the lines of the traffic light example, 
+there are many in our collection that are more realistic. 
+For instance, Anzu has controller specifications for a generalized IBM buffer and an AMBA bus. 
+LTLMoP and TuLiP have specifications for robot controllers that have been demonstrated on simulated and/or real robots. 
+And Salty has specifications for controllers to coordinate the actions of teams of vehicles that have been demonstrated in simulation.
 
-\figref{fig:timingResults} shows the amount of time needed to analyze examples as a function of number of transitions in the Moore machine representing the controller, with examples that could not be analyzed due to memory errors set to -10.
-Results were generated on a Linux VM given 24 GB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
-On most examples in which the Moore machine encoding the controller had less than 4000 transitions, 
+\figref{fig:timingResults} shows the amount of time needed to analyze examples as a function of total number of transitions 
+in the Moore machine representing the controller, 
+with examples that could not be analyzed due to memory errors set to 1. 
+Results were generated on a Linux VM given 24GB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
+We ran 40 examples in total. 
+Results for 33 examples are plotted. 
+On most examples with less than 4000 transitions, 
 SPARK was able to completely verify or prove the generated code complies with its specification. 
-Examples that had more than 4000 transitions required too much memory to analyze, resulting in errors when attempting to verify them in SPARK. 
+Examples that had more than 4000 transitions (the 7 unplotted examples) required too much memory to analyze, 
+resulting in errors when attempting to verify them in SPARK.
+
 Of examples with less than 4000 transitions, two resulted in errors. 
-These two examples had abnormally large specifications consisting of approximately 1000 terms each, whereas most other examples had 500 or less.
-Two could only be partially proven. 
+These two examples had abnormally large specifications consisting of approximately 1000 atomic propositions each, 
+whereas most other examples with a similar number of transitions had 500 or less. 
+Such cases occur, e.g. when systems include a large number of inputs and/or outputs but have very tight specifications on 
+how they can transition, leading to large specifications and therefore lengthy pre- and postconditions but relatively small controllers.
+
+Two examples with less than 4000 transitions could only be partially proven. 
 One was relatively large, with about 2000 transitions. 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{salty.eps}
-\caption{Timing results for example controllers in SPARK as a function of number of transitions in the Moore machine representing the controller.
-Examples marked ``proven'' were fully verified, examples marked ``partial'' were partially verified, and examples marked ``error'' were too big to analyze.}
+\includegraphics[width=\textwidth]{salty_log.eps}
+\caption{Timing results for example SPARK controllers as a function of number of transitions in the controller.
+``Proven'' examples were fully verified, ``partial'' examples were partially verified, and ``error'' examples were too big to analyze.}
 \label{fig:timingResults}
@@
 were derived from environment input values and used to look up the next \lstinline{State_Num} and corresponding system output values. 
-Ghost functions consisting of nested quantified expressions were used to check that in each state, initial and transition properties held 
-according to values stored in the current state's hashed map and the outputs of next states. 
+Ghost functions consisting of nested quantified expressions were used to check that in each state, specification properties held 
+using input and output values encoded by the current state and all states reachable as contained in the hashed maps.
 These functions comprised the post-condition of a function that initialized the controller's lookup table. 
@@
 While the use of formal containers was intuitive, they are more complex to reason about with respect to proof 
-because they require reasoning about models of the containers and the effects of container operations. 
+because they require reasoning about models of the containers. 
 Encoding the lookup table as a case statement is more straightforward, because for instance, 
@@
 
-Analysis with SPARK did reveal unexpected behaviors in two controllers synthesized from our corpus of examples. 
+Analysis with SPARK did reveal unexpected behaviors in two example controllers. 
 In particular, these controllers contained empty case statements in the \lstinline{Move} procedure. 
@@
 Upon inspection of the specifications, we found that these two examples\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} had 
-environment transition terms that allowed the environment to produce an input that would always cause 
+environment transition terms that allowed the environment to produce an input that would cause 
 the environment to violate its own specification on the next time step. 
-These specifications had the form $\always \lnot p$ instead of $\always \lnext \lnot p$. 
+To explain, suppose $p$ is an input. 
+Then these specifications had the form $\always \lnot p$ instead of $\always \lnext \lnot p$. 
 The distinction is important.
-In the former case, the environment can choose $p$ for its next input and not violate the specification on the current time step. 
-However, it violates it on the next time step when $p$ becomes the ``current'' value instead of the ``next'' value. 
-The latter case does not allow $p$ on the next time step, and so it also cannot hold on future ``current'' steps. 
-Generally, specifications should follow the latter form, 
+In GR(1), specifications are interpreted in the context of a two-player game in which transition specifications 
+are over the current state and the next state. 
+In the former case, the environment can choose $p$ for the ``next'' time step, and this does not violate $\always \lnot p$ 
+in the ``current'' time step, since the new ``current'' time step is only constructed once the environment and system have both taken their turn. 
+However, once the next state is reached, $p$ becomes the new ``current'' value, and $\always \lnot p$ 
+will now be violated no matter what the environment chooses.
+Generally, specifications should follow the latter form $\always \lnext \lnot p$, 
 and this was indeed an error in these specifications, so we fixed them.
-However, we also modified Salty to put \lstinline{raise Program_Error} for such cases and checked 
+However, we also modified Salty to \lstinline{raise Program_Error} for such cases and checked 
 that we were still able to prove $\varphi_e \limplies \varphi_s$ for these examples  
-(since reaching these cases would require violating the precondition $\varphi_e$ on the environment). 
+(since reaching these cases would require violating the precondition $\varphi_e$). 
+Such cases amount to instances in which the precondition on \lstinline{Move} is violated, which is why we allow the user to execute \lstinline{Env_Trans} 
+as discussed in \sectref{sec:implementation}. 
 We found that in other Salty language targets, procedures equivalent to \lstinline{Move} would have raised random runtime exceptions 
@@
 but they were inefficiently encoded. 
-We modified Salty to handle such cases by removing the environment, functions over the environment, and
+We plan to modify Salty to handle such cases by removing the environment, functions over the environment, and
 all references to the environment in all pre- and postconditions.
-This greatly decreased verification time and also reduced the size and increased the efficiency of the code.
+This greatly decreases verification time and also reduces the size and increases the efficiency of the code.
 
@@
 We were able to successfully use SPARK to verify safety and transition properties of moderately sized controllers 
-synthesized from GR(1) specifications by the tool Salty. 
-Encoding the controllers and all of the annotations necessary for these controllers to prove automatically was relatively straighforward. 
-The act of performing ``end-to-end'' verification with SPARK on such controllers was valuable because (1) it revealed implementation errors for certain 
+synthesized by Salty from GR(1) specifications. 
+Encoding the controllers and all of the annotations necessary for these controllers to prove automatically was relatively straighforward, 
+and it was satisfying to be able to generate proofs using a single tool rather than having to use mulitple tools to perform verification.
+Furthermore, the act of performing ``end-to-end'' verification with SPARK on such controllers was valuable because (1) it revealed implementation errors for certain 
 types of specifications that result in random runtime errors in other Salty target languages, and (2) it revealed cases in which controllers were inefficiently encoded, i.e.

### 2020-01-28 doc/2020-CAV/paper.tex
Minor edits

@@
 Its initial state is always the last possible state number. 
-Each input and output is of type \lstinline{Boolean}. 
+In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
 When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
@@
 The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
-\lstinline{State_To_Input_Mapping} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
-\lstinline{State_To_Input_Mapping} returns \lstinline{True} if and only if 
+\lstinline{State_To_Input_Mapping (C)} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
+\lstinline{State_To_Output_Mapping (C)} returns \lstinline{True} if and only if 
 \sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
@@
 implemented in the body (not shown), since they make use of input and output values stored in the \lstinline{Controller}, whose fields are private.
+
+The public procedure \lstinline{Move} (lines 18-23) transitions a controller based on its current state and 
+next set of input value(s), and it produces the next set of output value(s). 
+It has a precondition that if the controller is in its initial state, 
+inputs must satisfy $\varphi_e^i$; otherwise they must satisfy $\varphi_e^t$.
+The aspect \lstinline{Contract_Cases} specifies additional sets of preconditions paired with postconditions, 
+where the set of all preconditions must be mutually exclusive and exhaustively cover the entire input space.
+The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
+Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
+for the right-hand side, they refer to values after evaluation.
+Therefore, the aspect \lstinline{Old} can be used on the right to reference the value of a variable before evaluation of the subprogram. 
+Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
+the first set of output value(s) produced should satisfy $\varphi_s^i$ and the controller should no longer be in its initial state; 
+if it is not in its initial state, then the output value(s) produced should satisfy $\varphi_s^t$, 
+evaluated based on the most recent input and output values stored in \lstinline{C'Old}, the next input value(s) just provided (in this case in \lstinline{tick}), 
+and the next output values just generated (in this case in the record \lstinline{System}). 
+This set of contract cases embodies our main proof goal, i.e. verification of system initial and transition properties from the original GR(1) specification for the controller.
 
@@
 \end{figure}
-
-The public procedure \lstinline{Move} (lines 18-23) transitions a controller based on its current state and 
-next set of input value(s), and it produces the next set of output value(s). 
-It has a precondition that if the controller is in its initial state, 
-inputs must satisfy $\varphi_e^i$; otherwise they must satisfy $\varphi_e^t$.
-The aspect \lstinline{Contract_Cases} specifies additional sets of preconditions paired with postconditions, 
-where the set of all preconditions must be mutually exclusive and exhaustively cover the entire input space.
-The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
-Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
-for the right-hand side, they refer to values after evaluation.
-Therefore, the aspect \lstinline{Old} can be used on the right to reference the value of a variable before evaluation of the subprogram. 
-Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
-the first set of output value(s) produced should satisfy $\varphi_s^i$ and the controller should no longer be in its initial state; 
-if it is not in its initial state, then the output value(s) produced should satisfy $\varphi_s^t$, 
-evaluated based on the most recent input and output values stored in \lstinline{C'Old}, the next input value(s) just provided (in this case in \lstinline{tick}), 
-and the next output values just generated (in this case in the record \lstinline{System}). 
-This set of contract cases embodies our main proof goal, i.e. verification of system initial and transition properties from the original GR(1) specification for the controller.
 
@@
 \end{figure}
+
+\vspace{-1em}
 
@@
 Based on our first round of SPARK analysis, we also found some examples in our database that did not have any inputs, 
-i.e. they essentially amounted to synthesizing a system independent of an environment. 
+i.e. they amounted to synthesizing a system independent of an environment. 
 In those cases, we had specifications for a non-existent environment that were vacuous, 
 and this was causing SPARK to take an abnormally long amount of time to verify these controllers, given their relatively small size. 
-These controllers were did not have errors per se, 
+These controllers did not have errors per se, 
 but they were inefficiently encoded. 
@@
 Encoding the controllers and all of the annotations necessary for these controllers to prove automatically was relatively straighforward, 
-and it was satisfying to be able to generate proofs using a single tool rather than having to use mulitple tools to perform verification.
+and it was satisfying to be able to generate proofs using a single tool rather than having to use multiple tools to perform verification.
 Furthermore, the act of performing ``end-to-end'' verification with SPARK on such controllers was valuable because (1) it revealed implementation errors for certain 

### 2020-04-23 doc/2020-SEFM/paper.tex
Changed 2020-CAV directory to 2020-SEFM

--- 2020-01-28 doc/2020-CAV/paper.tex
@@
 %
-\title{End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
+\title{\large Experience Report: End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
 %
@@
 The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
-\lstinline{State_To_Input_Mapping (C)} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
-\lstinline{State_To_Output_Mapping (C)} returns \lstinline{True} if and only if 
+\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
+\lstinline{State_To_Output_Mapping(C)} returns \lstinline{True} if and only if 
 \sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 

### 2020-04-23 doc/2020-SEFM/paper_back_v3.tex
Changed 2020-CAV directory to 2020-SEFM

--- 2020-01-28 doc/2020-CAV/paper.tex
@@
 The result of synthesis is a \emph{control protocol} or \emph{strategy} that can be expressed as a Moore machine. 
-%Note that in the case of GR(1), synthesized Moore machines not only produce a single set of system output values in each state, 
-%there is a unique set of environment input values that brings the controller into each state.
+In a Moore machine, each state produces a unique set of system output values, 
+and in the case of GR(1), synthesized Moore machines have the additional property that 
+there is a unique set of environment input values that brings the system into each state.
 
@@
 For example, RATSY %(Requirements Analysis Tool with Synthesis) 
-\cite{bloem2010ratsy} has a focus on circuit design and synthesizes designs 
+\cite{bloem2010ratsy} has a focus on circuit design and can synthesize designs 
 encoded in BLIF, %(Berkeley Logic Interchange Format),
@@
 %(SmalL bUt Complete GROne Synthesizer) 
-\cite{Ehlers2016} is architected to allow users to tailor synthesis algorithms,
+\cite{Ehlers2016} is architected to allow users to easily tailor synthesis algorithms for specific applications,
  e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
@@
 tool implementation errors could still result in errors in synthesized controllers. 
-We are therefore interested in ``end-to-end'' verification of synthesized controllers. % with respect to their original GR(1) specifications. 
+We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications. 
 Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK. 
@@
 However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
-synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their GR(1) specifications in SPARK. 
+synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their functional specifications in SPARK. 
 To date, we have achieved this goal for a subset of properties comprising GR(1) specifications, 
@@
 In what follows, in \sectref{sec:implementation} we discuss the SPARK implementation. 
-In \sectref{sec:caseStudies}, we discuss our results on examples pulled from various sources.
+In \sectref{sec:caseStudies}, we demonstrate the efficacy of our approach on examples pulled from various sources.
 In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to structure synthesized controllers for proof, 
-examples in which SPARK revealed unexpected controller behaviors, and caveats related to GR(1) specifications. 
+examples in which SPARK revealed unexpected controller behaviors, and caveats related to the interpretation of GR(1) specifications. 
 We end with concluding remarks in \sectref{sec:conclusions}, including a discussion of future work.
@@
 
-To explain the structure of our synthesized SPARK controllers, consider a simple example of a traffic light controller. 
-Let the controller's single input be $tick$. 
-Color changes occur in every state in which $tick$ is true.
-The traffic light's color cycles infinitely over the sequence red $\rightarrow$ green $\rightarrow$ yellow $\rightarrow$ red $\rightarrow$ $\ldots$.
+To explain the structure of the controllers we generate in SPARK, consider a simple example of a controller for a traffic light. 
+Color changes are triggered in every state in which an input signal's value is true.
+Let this input be called $tick$.
+The traffic light's color changes from red to green to yellow to red $\ldots$, repeating the cycle infinitely. 
 Let the output variables then be $red$, $yellow$, and $green$.
@@
 and the color should remain the same whenever $tick$ is false \eqnref{eq:same}.
-
-\renewcommand{\floatpagefraction}{.8}%
 
@@
 
-In SPARK, subunits consist of a specification and a body, 
-which Salty generates. 
-% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
+\sloppy{
+In SPARK, subunits consist of a specification and a body. 
+For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
 The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
-Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
-In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
-% (possibly violating the controller's GR(1) specification). 
+Type \lstinline{Controller} encodes a Moore machine suitable for representing the state of a synthesized controller. 
+In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state 
+(possibly violating the controller's GR(1) specification). 
 In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
@@
 there is exactly one set of input values that brought the controller into that state. 
-Therefore, \lstinline{Controller} includes a type invariant (lines 36-37) that captures this property, 
-where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} state whether the controller has the corresponding state number and input or output value(s), respectively. 
-% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
-Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
+Therefore, \lstinline{Controller} includes a type invariant (lines 38-39) that captures this property, 
+where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
+encode lookup tables specifying the input and output value(s) for each state, respectively.
+Both of these are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
 i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
@@
 In this case, removing the ghost code can save significant memory, 
-since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
-internally encode large lookup tables that have to store information on every state of the controller, 
-which is often thousands of states or even millions of states for the largest example in our database.
-The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
-\lstinline{State_To_Input_Mapping} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
-\lstinline{State_To_Input_Mapping} returns \lstinline{True} if and only if 
-\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
-since \lstinline{tick = False} brings the system into state 1 and $red$ is true and $yellow$ and $green$ are false in state 1.
+especially on bigger examples where the tables encoded by functions  \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
+would have to store information on thousands of states.
+The logic for these functions is given in the body (not shown here), but for instance in state 1, \lstinline{State_To_Input_Mapping} returns \lstinline{False}, 
+since \lstinline{tick = False} as the system enters state 1, and \lstinline{State_To_Output_Mapping} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
+since $red$ is True and $yellow$ and $green$ are false as the system enters state 1.
+ }
 
@@
 The public function \lstinline{Env_Init} (line 9) checks whether input(s) satisfy $\varphi_e^i$. 
-It is implemented as an expression function, i.e. the implementation is given directly in the specification, 
+It is implemented as an expression function, i.e. the implementation is given directly in the specification and not the body, 
 because all terms needed to define it are visible in the public part of the specification. 
@@
 The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
-Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
-for the right-hand side, they refer to values after evaluation.
-Therefore, the aspect \lstinline{Old} can be used on the right to reference the value of a variable before evaluation of the subprogram. 
+Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to variable values before evaluation of the subprogram; 
+for the right-hand side, they refer to variable values after evaluation.
+For the right-hand side, the aspect \lstinline{Old} can therefore be used to reference the value of a variable before evaluation of the subprogram. 
 Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
@@
 that would not be allowed due to $\varphi_e^i$ or $\varphi_e^t$. 
-In the traffic light example, these are unnecessary because all possible combinations of input values are 
+In the traffic light example, they are unnecessary because all possible combinations of input values are 
 allowed out of each state. 
-In any case, SPARK will prove that such cases are not reachable.
+In any case, SPARK will prove that these cases are not reachable.
 
@@
 
-\figref{fig:timingResults} shows the amount of time needed to analyze examples as a function of number of transitions in the Moore machine representing the controller, with examples that could not be analyzed due to memory errors set to -10.
-Results were generated on a Linux VM given 24 GB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
+\figref{fig:timingResults} shows the amount of time needed to analyze examples as 
+a function of number of total transitions in the Moore machine representing the controller.
+
+approximate number of states, 
+with examples with unproved functional contracts marked in red and examples that could not be analyzed were set to -10 in red.
+Results were generated on a Linux VM given 24GB RAM and 4 processors on a MacBook Pro with a 2.9 GHz Intel Core i9 with 32GB RAM.
 On most examples in which the Moore machine encoding the controller had less than 4000 transitions, 
@@
 were derived from environment input values and used to look up the next \lstinline{State_Num} and corresponding system output values. 
-Ghost functions consisting of nested quantified expressions were used to check that in each state, initial and transition properties held 
+Ghost functions consisting of nested quantified expressions were used to assert that in each state, initial and transition properties held 
 according to values stored in the current state's hashed map and the outputs of next states. 

### 2020-04-23 doc/2020-SEFM/paper_joffrey.tex
Changed 2020-CAV directory to 2020-SEFM

--- 2020-01-28 doc/2020-CAV/paper.tex
@@
 Its initial state is always the last possible state number. 
-In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
+In this example, each input and output is of type \lstinline{Boolean}. 
 When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
@@
 
-The public procedure \lstinline{Move} (lines 18-23) transitions a controller based on its current state and 
-next set of input value(s), and it produces the next set of output value(s). 
-It has a precondition that if the controller is in its initial state, 
-inputs must satisfy $\varphi_e^i$; otherwise they must satisfy $\varphi_e^t$.
-The aspect \lstinline{Contract_Cases} specifies additional sets of preconditions paired with postconditions, 
-where the set of all preconditions must be mutually exclusive and exhaustively cover the entire input space.
-The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
-Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
-for the right-hand side, they refer to values after evaluation.
-Therefore, the aspect \lstinline{Old} can be used on the right to reference the value of a variable before evaluation of the subprogram. 
-Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
-the first set of output value(s) produced should satisfy $\varphi_s^i$ and the controller should no longer be in its initial state; 
-if it is not in its initial state, then the output value(s) produced should satisfy $\varphi_s^t$, 
-evaluated based on the most recent input and output values stored in \lstinline{C'Old}, the next input value(s) just provided (in this case in \lstinline{tick}), 
-and the next output values just generated (in this case in the record \lstinline{System}). 
-This set of contract cases embodies our main proof goal, i.e. verification of system initial and transition properties from the original GR(1) specification for the controller.
-
 \begin{figure}
@@
 \end{figure}
+
+The public procedure \lstinline{Move} (lines 18-23) transitions a controller based on its current state and 
+next set of input value(s), and it produces the next set of output value(s). 
+It has a precondition that if the controller is in its initial state, 
+inputs must satisfy $\varphi_e^i$; otherwise they must satisfy $\varphi_e^t$.
+The aspect \lstinline{Contract_Cases} specifies additional sets of preconditions paired with postconditions, 
+where the set of all preconditions must be mutually exclusive and exhaustively cover the entire input space.
+The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
+Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
+for the right-hand side, they refer to values after evaluation.
+Therefore, the aspect \lstinline{Old} can be used on the right to reference the value of a variable before evaluation of the subprogram. 
+Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
+the first set of output value(s) produced should satisfy $\varphi_s^i$ and the controller should no longer be in its initial state; 
+if it is not in its initial state, then the output value(s) produced should satisfy $\varphi_s^t$, 
+evaluated based on the most recent input and output values stored in \lstinline{C'Old}, the next input value(s) just provided (in this case in \lstinline{tick}), 
+and the next output values just generated (in this case in the record \lstinline{System}). 
+This set of contract cases embodies our main proof goal, i.e. verification of system initial and transition properties from the original GR(1) specification for the controller.
 
@@
 \end{figure}
-
-\vspace{-1em}
 
@@
 Based on our first round of SPARK analysis, we also found some examples in our database that did not have any inputs, 
-i.e. they amounted to synthesizing a system independent of an environment. 
+i.e. they essentially amounted to synthesizing a system independent of an environment. 
 In those cases, we had specifications for a non-existent environment that were vacuous, 
 and this was causing SPARK to take an abnormally long amount of time to verify these controllers, given their relatively small size. 
-These controllers did not have errors per se, 
+These controllers were did not have errors per se, 
 but they were inefficiently encoded. 

### 2020-04-28 doc/2020-SEFM/paper.tex
Paper edits for SEFM

@@
 % \renewcommand\UrlFont{\color{blue}\rmfamily}
+\allowdisplaybreaks
 
@@
 %
-\title{\large Experience Report: End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
+\title{\large End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK}
+
+%\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
 %
@@
 %
-%\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
-%James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
-%Joffrey Huguet\inst{3}\orcidID{2222--3333-4444-5555}}
-\author{Laura Humphrey\inst{1} \and
-James Hamil\inst{2} \and
-Joffrey Huguet\inst{3}}
+\author{Author 1\inst{1} \and
+Author 2\inst{2} \and
+Author 3\inst{3}}
+
+%\author{Laura Humphrey\inst{1} \and
+%James Hamil\inst{2} \and
+%Joffrey Huguet\inst{3}}
 %
-\authorrunning{L. Humphrey et al.}
+
+% \authorrunning{L. Humphrey et al.}
+\authorrunning{Author 1 et al.}
+
 % First names are abbreviated in the running head.
@@
 %
-\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
-\email{laura.humphrey@us.af.mil} \and
-LinQuest Corp., Beavercreek, OH 45431, USA \\
-\email{james.hamil.ctr@us.af.mil}  \and
-AdaCore, F-75009 Paris, France \\
- \email{huguet@adacore.com}}
+\institute{Institute 1 \\
+\email{author1@email.com} \and
+Institute 2 \\
+\email{author2@email.com}  \and
+Institute 3 \\
+ \email{author3@email.com}}
+%
+%\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
+%\email{laura.humphrey@us.af.mil} \and
+%LinQuest Corp., Beavercreek, OH 45431, USA \\
+%\email{james.hamil.ctr@us.af.mil}  \and
+%AdaCore, F-75009 Paris, France \\
+% \email{huguet@adacore.com}}
 %
@@
  e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
+ %% -- TODO: put more details about input types that Salty allows? About what optimizations it does?
 Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
-a back-end that turns controller designs into executable software implementations in a variety of programming languages.
+a back-end that turns controller designs into executable software implementations in a variety of programming languages, 
+including Python, C++, Java, and now SPARK.
 
@@
 SPARK is both a programming language with a specification language and associated verification toolset \cite{hoang2015spark}. 
+SPARK translates code to the Why language for processing by Why3, 
+and Why3 uses other automated provers -- including Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
 Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers by
@@
 i.e. system initial and transition properties, for moderately-sized controllers.
-In what follows, in \sectref{sec:implementation} we discuss the SPARK implementation. 
-In \sectref{sec:caseStudies}, we discuss our results on examples pulled from various sources.
-In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to structure synthesized controllers for proof, 
-examples in which SPARK revealed unexpected controller behaviors, and caveats related to GR(1) specifications. 
+
+In what follows, in \sectref{sec:implementation} we describe the structure of synthesized controller implementations in SPARK 
+by walking through a simple example involving a traffic light. 
+To evaluate whether verification of synthesized SPARK controllers is feasible and has utility, 
+we pulled in a large body of examples from various sources. 
+In \sectref{sec:vipExample}, we take a detailed look at one of these examples, a controller that coordinates the actions of 
+a team of unmanned air vehicles (UAVs) performing an escort mission, 
+and we describe how SPARK revealed an error in the controller's specification. 
+In \sectref{sec:caseStudies}, we give an overview of results for the rest of the examples, with a focus on scalability. 
+In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to structure synthesized controllers for proof.  
+% examples in which SPARK revealed unexpected controller behaviors, and caveats related to GR(1) specifications. 
 We end with concluding remarks in \sectref{sec:conclusions}, including a discussion of future work.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Implementation and Verification in SPARK}
+\section{Implementation and Verification of Synthesized Controllers in SPARK}
 \label{sec:implementation}
 
-To explain the structure of our synthesized SPARK controllers, consider a simple example of a traffic light controller. 
+We have extended the open source tool Salty\footnote{https://github.com/GaloisInc/salty/} to produce software implementations of controllers synthesized from GR(1) specifications in SPARK. 
+Given a GR(1) specification in Salty format, Salty does some preprocessing to sanity check, optimize, and translate the specification to Slugs format. 
+Salty then calls Slugs to synthesize a controller design from the processed specification. 
+Slugs provides the option of returning the controller design as a Moore machine expressed in text format, 
+which is what Salty uses to create an executable software implementation of the controller. 
+This Moore machine encodes all the states of the controller, including a single initial state. 
+It also encodes the transition relation between states, i.e. the controller's next state given the current state 
+and the next set of input values. 
+Note that this transition relation is deterministic, i.e. there is at most one possible next state for each state and set of input values. 
+Every state then encodes a set of output values to be produced each time the controller transitions to that state. 
+As in all Moore machines, each state produces exactly one set of output values. 
+Controllers synthesized from GR(1) specifications have the additional property that for each state, 
+the set of input values that brings the controller into that state is the same, regardless of what the previous state was.
+% there is exactly one set of input values that brings the controller into that state.
+
+If SPARK is chosen as the target progamming language, Salty translates this textual Moore machine representation 
+of the controller to an implementation in SPARK. 
+To give a brief overview, the Moore machine representing the controller is encoded as a record (analogous to a struct in C) 
+that stores the controller's current state and a copy of the current input and output values, 
+i.e. the inputs and outputs associated with the most recent transition. 
+There is then a ``move'' procedure that implements the transition relation. 
+This procedure changes the controller's state based on the next set of input values, produces the next set of output values, 
+and updates its internal copy of current input and output values based on these most recent input and output values. 
+Salty also synthesizes all of the annotations necessary to encode the specifications of interest 
+and verify that the controller satisfies those specifications. 
+In terms of specifications, we seek to prove that the controller satisfies the system initial properties $\varphi^s_i$ 
+and system transition properties $\varphi^s_t$. 
+Recall that GR(1) specifications take the form 
+$(\varphi^e_i \land \varphi^e_t \land \varphi^e_l)  \limplies (\varphi^s_i \land \varphi^s_t \land \varphi^s_l)$. 
+Note however that for controllers synthesized from GR(1) specifications,
+system initial and transition properties should hold regardless of whether environment liveness properties hold. 
+We are therefore interested in showing that 
+$(\varphi^e_i \land \varphi^e_t)  \limplies (\varphi^s_i \land \varphi^s_t)$. 
+For the SPARK implementation, Salty therefore generates functions corresponding to $\varphi^e_i$ and $\varphi^e_t$,
+which are used as a precondition for the move procedure. 
+It also generates functions corresponding to $\varphi^s_i$ and $\varphi^s_t$, 
+which are used as a postcondition for the move procedure. 
+If SPARK can verify that the move function satisfies the postcondition given the precondition, 
+this verifies that $(\varphi^e_i \land \varphi^e_t)  \limplies (\varphi^s_i \land \varphi^s_t)$. 
+In terms of additional annotations needed to prove this property in SPARK, Salty generates 
+``state to input mapping'' and ``state to output mapping'' functions that encode which input values 
+and output values are associated with each state; these are used to define an invariant for the controller.  
+This is essentially all this is required to prove system initial and system transition properties of synthesized controllers in SPARK.
+
+To explain synthesized SPARK controllers and annotations in more detail, we consider a simple example involving a traffic light controller. 
 Let the controller's single input be $tick$. 
@@
 
-In SPARK, subunits consist of a specification and a body, 
-which Salty generates. 
+In SPARK, subunits consist of a specification and a body.
 % For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
@@
 In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
-current input value(s), and the current output value(s). 
+current input value(s), and current output value(s). 
 Its initial state is always the last possible state number. 
@@
 there is exactly one set of output values produced in that state. 
-Furthermore, controllers synthesized from GR(1) specifications have the additional property that 
+Furthermore, recall that controllers synthesized from GR(1) specifications have the additional property that 
 there is exactly one set of input values that brought the controller into that state. 
@@
 i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
-unless it is specified with a certain switch during compilation that they should be.
+unless it is specified during compilation that they should be.
 Executing ghost code is often used for debugging, e.g. to check whether an unproved property is actually true. 
@@
 The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
-\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False} and 
+\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False}, and 
 \lstinline{State_To_Output_Mapping(C)} returns \lstinline{True} if and only if 
@@
 because all terms needed to define it are visible in the public part of the specification. 
-In this example, since $\varphi_e^i = \top$ as pulled from the Salty specification, this function always simply returns \lstinline{True}. 
+In this example, since $\varphi_e^i = \top$ in the Salty specification, this automatically generated function always simply returns \lstinline{True}. 
 The public function \lstinline{Sys_Init} (lines 10-11) checks whether outputs(s) satisfy $\varphi_s^i$. 
@@
 We briefly note that Salty includes language features that can result in different constructs being used 
-to represent inputs and outputs in synthesized controllers, 
-including enumerations and integers. 
-For instance, enumerations encode that an enumerated input or output can have exactly one of a set of values, 
+to represent inputs and outputs in synthesized controllers, including enumerations and integers.
+For instance, enumerations encode that an enumerated input or output can have exactly one of a set of values at a time, 
 as in the traffic light being exactly one color as expressed in $\varphi_s^i$ and part \eqnref{eq:mutex} of $\varphi_s^t$. 
 In such cases, enumerations or integers can make both specifications and code more compact and easier to read and understand. 
+As a technical aside, during synthesis, enumerations and integers are translated to a bit vector representation along with the necessary 
+initial and transition specifications to encode properties such as values being mutually exclusive and rules for addition and subtraction over integers. 
+Once synthesis is complete, values are translated back to their original enumeration or integer representation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% IMPLEMENTATION and Verification in SPARK
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Example Controller for a Team of UAVs}
+\label{sec:vipExample}
+
+In order to evaluate the feasibility and utility of our approach, we collected GR(1) specifications from a variety of sources. 
+This includes the Salty repository, which contains examples of GR(1) specifications for control of UAV teams performing different missions. 
+One of these encodes the rules for a ``VIP Escort'' mission, 
+in which one UAV is designated as a ``very important person'' (VIP) that must always be ``escorted'' by a friendly surveillance UAV when it moves
+and protected from an ``enemy'' UAV. 
+The VIP and surveillance UAVs are controlled by the system, while the enemy UAV is controlled by the environment.
+The mission map contains regions that the UAVs can move between.  
+``Escorting'' consists of ensuring that 1) the VIP only enters regions previously visited by a surveillance UAV 
+and 2) whenever the VIP changes regions, a surveillance UAV ``tracks'' it, i.e. moves with it between regions. 
+``Protection'' consists of ensuring that the VIP is never in the same region as the enemy UAV. 
+Additional rules include constraints and liveness requirements on how the UAVs can move between regions and which regions they start in.
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{smallMission}
+  \caption{A VIP escort multi-UAV mission as simulated in OpenAMASE using a Salty-synthesized controller that issues commands to the UAVs through OpenUxAS.}
+  \label{fig:vipEscort}
+\end{figure}
+
+A particular instantiation of this mission is shown in \figref{fig:vipEscort}, as depicted in 
+the open source Aerospace Multi-Agent Simulation Environment (OpenAMASE)\footnote{https://github.com/afrl-rq/OpenAMASE}. 
+It includes the VIP, two surveillance UAVs numbered 1 and 2, one enemy UAV, and five regions numbered 1 to 5. 
+Environment inputs include integer variable $loc_e \in \{1 \dots 5\}$, which encodes the current region of the enemy UAV, 
+and Boolean variables $sr_i$ for $i = \{1 \ldots 5\}$, where $sr_i$ is true if and only if region $i$ has been visited at some point by a surveillance UAV. 
+% Note that here, we assume surveillance of a region occurs if and only if a surveillance UAV has been in that region. 
+System outputs include integer variables $loc_v, loc_{s1}, loc_{s2} \in \{1 \dots 5\}$, which indicate the current region of the 
+VIP and surveillance UAVs 1 and 2, and Boolean variables $vTrack_1$ and $vTrack_2$, which indicate whether surveillance UAVs 1 and 2 
+are executing a behavior to follow the VIP. 
+Note that low-level control, e.g. waypoint planning and sensor steering, 
+is implemented by the open source Unmanned Systems Autonomy Services (OpenUxAS)\footnote{https://github.com/afrl-rq/OpenUxAS}. 
+The controller synthesized by Salty implements high-level decision logic, 
+and OpenUxAS monitors the state of the controller and translates its current output values to a set of UAV commands that are simulated in OpenAMASE.
+For example, when $loc_v$ changes value to 2, OpenUxAS commands the VIP to follow a path from its current region to region 2, 
+or when $vTrack_1$ changes from false to true, OpenUxAS commands surveillance UAV 1 to fly next to the VIP. 
+The high-level controller synthesized by Salty makes some assumptions about the behaviors implemented by OpenUxAS, 
+in this case, mainly that all UAVs move at the same speed and transition to new regions at the same time. 
+A workflow for connecting Salty-implemented controllers with OpenUxAS and OpenAmase is described in \cite{elliott2019salty}, 
+and all of the scripts, configuration files, etc. needed to run this example are available on the Salty repository.
+
+%Note that all UAVs are assumed to move synchronously, 
+%i.e. if any change regions, the change happens at the same time.
+
+For this mission, the GR(1) specifications for the environment are:
+%
+{\small
+\begin{align}
+\varphi_e^i  = & \;\; (loc_e = 4) \land \lnot sr_1 \land \lnot sr_2 \land sr_3 \land \lnot sr_4 \land sr_5  \label{eq:vipEnvInit} \\
+\varphi_e^t  = & \bigwedge_{i = \{1 \ldots 5\}} \always \Big((loc_{s1} = i) \lor  (loc_{s2} = i) \limplies \lnext sr_i \Big) \land \mbox{} \label{eq:srSet} \\
+ & \bigwedge_{i = \{1 \ldots 5\}} \always \Big(\big(\lnot (loc_{s1} = i) \land  \lnot (loc_{s2} = i) \land \lnot sr_i \limplies \lnext \lnot sr_i\big) \land \mbox{} \label{eq:vipEnvNoSp} \\
+ & \quad\quad\quad\quad\quad \big(sr_i \limplies \lnext sr_i\big)\Big) \land \mbox{} \label{eq:vipEnvSetSp} \\
+% &  \bigwedge_{i = \{1 \ldots 5\}} \always  \big(sr_i \limplies \lnext sr_i\big) \land \mbox{} \label{eq:vipEnvSetSp} \\
+ & \;\; \always \lnot (loc_e = 1) \land \always \lnot (loc_e = 2) \label{eq:vipEnvDisallowedRegions} \\
+\varphi_e^l = & \;\; \always \eventually \lnot (loc_e = 3) \land \always \eventually \lnot (loc_e = 4) \land \always \eventually \lnot (loc_e = 5). \label{eq:vipEnvLiveness}
+\end{align}
+}
+\noindent These express \eqnref{eq:vipEnvInit} that the enemy UAV starts in region 4; 
+\eqnref{eq:srSet} that a region is considered to be surveilled after either one of the surveillance UAVs is in it; 
+\eqnref{eq:vipEnvNoSp} that a previously unsurveilled region remains unsurveilled if neither surveillance UAV is in it; 
+\eqnref{eq:vipEnvSetSp} that once a region is surveilled, it remains surveilled; 
+\eqnref{eq:vipEnvDisallowedRegions} that the enemy UAV cannot go to regions 1 or 2; and
+\eqnref{eq:vipEnvLiveness} that the enemy UAV must infinitely often go to regions 3, 4, and 5.
+
+GR(1) specifications for the system are:
+{\small
+\begin{align}
+\varphi_s^i  = & \;\; (loc_v = 2) \land (loc_{s1} = 3) \land (loc_{s2} = 5) \land \lnot vTrack_1 \land \lnot vTrack_2 \label{eq:vipSysInit} \\
+\varphi_s^t  =  & \quad\quad\quad\;\; \always \Big( \lnot(loc_v = \lnext loc_v) \limplies (\lnext vTrack_1 \lor \lnext vTrack_2) \Big)  \land \mbox{} \label{eq:vipNoMoveUnlessTracked} \\
+ & \bigwedge_{i = \{1 \ldots 2\}} \always \Big(vTrack_i \limplies (sloc_i = loc_v)\Big) \land \mbox{} \label{eq:vipSysVtrack} \\
+& \bigwedge_{i = \{1 \ldots 5\}} \always \Big((loc_v = i) \limplies \lnot(loc_e = i)\Big) \land \mbox{} \label{eq:vipSysNoEloc} \\
+ & \bigwedge_{i = \{v, s1, s2\}} \always \Big(\big( \lnext(loc_i = 1) \limplies (loc_i = 1) \lor (loc_i = 2) \lor (loc_i = 3)   \big) \land \mbox{} \label{eq:vipSysTrans1} \\
+ & \quad\quad\quad\quad\quad\;\;\; \big(\lnext (loc_i = 2) \limplies (loc_i = 1) \lor (loc_i = 2) \lor (loc_i = 3)   \big) \land \mbox{} \label{eq:vipSysTrans2} \\
+  & \quad\quad\quad\quad\quad\;\;\; \big(\lnext (loc_i = 3) \limplies \bigvee_{j=\{1\ldots5\}} (loc_i = j) \big) \land \mbox{}  \label{eq:vipSysTrans3} \\
+%  & \quad\quad\quad\quad\quad\;\;\; \big((loc_i = 3) \limplies (loc_i = 1) \lor (loc_i = 2) \lor (loc_i = 3) \lor \mbox{} \nonumber \\ 
+%  & \quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad     (loc_i = 4) \lor (loc_i = 5)   \big) \land \mbox{} \nonumber \\
+  & \quad\quad\quad\quad\quad\;\;\; \big(\lnext (loc_i = 4) \limplies (loc_i = 3) \lor (loc_i = 4) \lor (loc_i = 5)   \big) \land \mbox{} \label{eq:vipSysTrans4} \\
+  & \quad\quad\quad\quad\quad\;\;\; \big(\lnext (loc_i = 5) \limplies (loc_i = 3) \lor (loc_i = 4) \lor (loc_i = 5)  \big)  \Big) \land \mbox{} \label{eq:vipSysTrans5} \\
+\varphi_s^l = & \;\; \always \eventually (loc_v = 1) \land \always \eventually (loc_v = 5). \label{eq:vipSysLiveness}
+\end{align}
+}
+\noindent These express \eqnref{eq:vipSysInit} that the VIP starts in region 2, surveillance UAV 1 in region 3, and surveillance UAV 2 in region 5; 
+\eqnref{eq:vipNoMoveUnlessTracked} that the VIP does not change regions unless a surveillance UAV is tracking it; 
+\eqnref{eq:vipSysVtrack} that a surveillance UAV can only track the VIP if they are in the same region at the same time; 
+\eqnref{eq:vipSysNoEloc} that the VIP cannot be in the same region as the enemy UAV at the same time; 
+\eqnref{eq:vipSysTrans1} and \eqnref{eq:vipSysTrans2} that the VIP and surveillance UAVs can move from regions 1 or 2 to regions 1, 2, or 3;
+\eqnref{eq:vipSysTrans3} that the VIP and surveillance UAVs can move from region 3 to any other region; 
+\eqnref{eq:vipSysTrans4} and \eqnref{eq:vipSysTrans5} that the VIP and surveillance UAVs can move from regions 4 or 5 to regions 3, 4, or 5; 
+and \eqnref{eq:vipSysLiveness} that the VIP must go to regions 1 and 5 infinitely often.
+
+We have chosen to describe this particular example in detail because during the process of extending Salty to produce SPARK translations, 
+analysis with SPARK revealed a previously undetected problem with this example's specification. 
+As written, the specification is synthesizable and produces what appears at a glance to be a reasonable controller with 97 states. 
+In fact, we had previously run this example with OpenAMASE, OpenUxAS, and a Python controller synthesized by Salty. 
+However, we did not notice that 34 of the controller's 97 states do not actually have successors. 
+In SPARK, this resulted in empty case statements in the \lstinline{Move} procedure. 
+For example, for state 2 the case alternative statement is simply \lstinline{when 2 =>}, i.e. there are no other statements associated with this case. 
+Attempting to execute this case alternative would result in a runtime error, which SPARK detected. 
+In Python, this error went undetected; 
+the transition relation for the controller is encoded as a map, and entries corresponding to states without successors 
+simply had an empty list of ``next state'' values. 
+If the system had attempted to transition states based on an empty list, the result would have been an unhandled runtime error.
+
+The reason for this has to do with a subtlety the semantics of GR(1) specifications. 
+Recall that GR(1) specifications take the form $\varphi_e \limplies \varphi_s$. 
+Obviously a specification of this form is satisfied if $\varphi_e$ and $\varphi_s$ are both true, 
+but it is also satisfied if $\varphi_e$ is false. 
+Also recall from the introduction that GR(1) specifications are interpreted in the context of a two-player game 
+in which the environment takes its turn first and the system takes its turn second. 
+The issue here is that the environment is able to take transitions that will necessarily cause it to violate $\varphi_e^t$. 
+Note that $\varphi_e^t$ contains terms of the form $\always \lnot p$, specifically $\always \lnot (loc_e = i)$ for $i = \{1, 2\}$  \eqnref{eq:vipEnvDisallowedRegions}. 
+Note that a term of the form $\always \lnot p$ is not the same as a term of the form $\always \lnext \lnot p$. 
+The distinction is important. 
+In the former case, the environment can choose $p$ for the ``next'' time step, and this does not violate $\always \lnot p$ 
+in the ``current'' time step, since the new ``current'' time step is only constructed once the environment and system have both taken their turn. 
+However, once the next state is reached, $p$ becomes the new ``current'' value, and $\always \lnot p$ 
+will now be violated no matter what the environment chooses.
+Generally, specifications should follow the latter form $\always \lnext \lnot p$, 
+and this was indeed an error in these specifications, so we changed $\always \lnot (loc_e = 1) \land \always \lnot (loc_e = 2)$
+ to $\always \lnext \lnot (loc_e = 1) \land \always \lnext \lnot (loc_e = 2)$ in \eqnref{eq:vipEnvDisallowedRegions}.
+However, we also modified Salty to \lstinline{raise Program_Error} for such cases and checked 
+that we were still able to prove $\varphi_e \limplies \varphi_s$ for these examples  
+(since reaching these cases would require violating the precondition $\varphi_e$). 
+Such cases amount to instances in which the precondition on \lstinline{Move} is violated, which is why we allow the user to execute \lstinline{Env_Trans} 
+as discussed in \sectref{sec:implementation}. 
+We also plan to have Salty issue a warning when there are states with no successors, 
+as such cases are likely unintended by the user.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Case Studies}
+\section{Results}
 \label{sec:caseStudies}
 
-To check whether our approach has the potential to produce SPARK controllers that can be automatically verified to satisfy their original system initial and transition specifications, 
-we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, and Salty, all of which make their examples publicly available for download. 
+To further evaluate the utility and feasibility of our approach, we pulled additional example GR(1) specifications from a variety of sources including 
+Anzu, LTLMoP, TuLiP, Slugs, and Salty, all of which make their examples publicly available for download. 
 These examples -- including GR(1) specifications in Salty format, synthesized SPARK packages, and SPARK analysis results -- are all available on the 
-Salty GitHub repository\footnote{https://github.com/GaloisInc/salty/}, including the traffic light example of the previous section. 
+Salty GitHub repository, including the traffic light example of the previous section. 
 We note that while some examples are small and simple, e.g. demos along the lines of the traffic light example, 
@@
 
-Analysis with SPARK did reveal unexpected behaviors in two example controllers. 
-In particular, these controllers contained empty case statements in the \lstinline{Move} procedure. 
-Upon inspection of the Slugs output, these corresponded to states with no successors. 
-Upon inspection of the specifications, we found that these two examples\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} had 
-environment transition terms that allowed the environment to produce an input that would cause 
-the environment to violate its own specification on the next time step. 
-To explain, suppose $p$ is an input. 
-Then these specifications had the form $\always \lnot p$ instead of $\always \lnext \lnot p$. 
-The distinction is important.
-In GR(1), specifications are interpreted in the context of a two-player game in which transition specifications 
-are over the current state and the next state. 
-In the former case, the environment can choose $p$ for the ``next'' time step, and this does not violate $\always \lnot p$ 
-in the ``current'' time step, since the new ``current'' time step is only constructed once the environment and system have both taken their turn. 
-However, once the next state is reached, $p$ becomes the new ``current'' value, and $\always \lnot p$ 
-will now be violated no matter what the environment chooses.
-Generally, specifications should follow the latter form $\always \lnext \lnot p$, 
-and this was indeed an error in these specifications, so we fixed them.
-However, we also modified Salty to \lstinline{raise Program_Error} for such cases and checked 
-that we were still able to prove $\varphi_e \limplies \varphi_s$ for these examples  
-(since reaching these cases would require violating the precondition $\varphi_e$). 
-Such cases amount to instances in which the precondition on \lstinline{Move} is violated, which is why we allow the user to execute \lstinline{Env_Trans} 
-as discussed in \sectref{sec:implementation}. 
-We found that in other Salty language targets, procedures equivalent to \lstinline{Move} would have raised random runtime exceptions 
-in such cases due to empty entries in their lookup tables.
+Analysis with SPARK did reveal some unexpected behaviors. 
+As discussed in \sectref{sec:vipExample}, there were two example controllers\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} 
+with specifications that resulted in states with no successors. 
+As a result, these controllers contained empty case statements in the \lstinline{Move} procedure. 
+We had previously run the Salty-synthesized controller in Python for the example in \sectref{sec:vipExample} and not noticed the error, 
+though it would have resulted in an unhandled runtime exception if one of the states without successors had been reached.
 
-Based on our first round of SPARK analysis, we also found some examples in our database that did not have any inputs, 
+A meta-analysis of SPARK timing results also revealed that other examples in our database that did not have any inputs, 
 i.e. they amounted to synthesizing a system independent of an environment. 

### 2020-05-27 doc/2020-SEFM/paper.tex
More details on SPARK in the Introduction

@@
 including Python, C++, Java, and now SPARK. 
-
 Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,''
 tool implementation errors could still result in errors in synthesized controllers.
-We are therefore interested in ``end-to-end'' verification of synthesized controllers. % with respect to their original GR(1) specifications. 
+We are therefore interested in ``end-to-end'' verification of synthesized controllers.
 Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK.
-SPARK is both a programming language with a specification language and associated verification toolset \cite{hoang2015spark}. 
-SPARK translates code to the Why language for processing by Why3, 
-and Why3 uses other automated provers -- including Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
-Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers by
-annotating the code, e.g. with assertions and loop invariants. 
-However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
-synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their GR(1) specifications in SPARK. 
-To date, we have achieved this goal for a subset of properties comprising GR(1) specifications, 
-i.e. system initial and transition properties, for moderately-sized controllers.
+
+SPARK, which is based on the Ada programming language, is both a programming language 
+with a specification language and an associated verification toolset \cite{hoang2015spark}.
+SPARK translates code to the Why language for processing by the Why3 theorem prover, 
+and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
+Though SPARK aims to perform fully automated verification, manual adaptation of the source code is in general necessary. 
+For example, if one were to translate source code originally written in Ada to SPARK, 
+one would have to remove unsupported features such as functions with side-effects, aliased variable names, exception handlers, etc.  
+Formal verification with SPARK also requires supplying requirements at the level of the source code in the 
+form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by indvidual subprograms.
+There are different levels to which one can use SPARK to verify a program. 
+Generally, the level of verification performed during development is incremental, going from lowest to highest. 
+Verification at the lowest level, ``stone level,'' is achieved when code is accepted by SPARK, 
+since the SPARK language has stricter legality rules than Ada. 
+Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
+flow analysis checks for errors such as reads of unintialized data or violations of user-specified data flow contracts. 
+Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
+e.g. division by zero and numeric overflow/underflow. 
+Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
+type invariants and subprogram preconditions and postconditions; 
+however, at this level, the specifications only partially specify the desired behavior of the code.  
+Verification at the next level, ``platinum level,'' consists verifying a complete set of specifications.
+At the moment, the SPARK code generated by Salty proves at gold level.  
+SPARK is able to prove the system initial and system transition property portions of the original GR(1) specification. 
+Proof of system transition properties in turn requires proving a type invariant on the underlying Moore machine representation of the controller; 
+this type invariant encodes that each state is reached by a specific set of input values and produces a specific set of output values. 
+The platinum level would be reached if the liveness properties were expressed and proved, 
+which we leave for future work.
 
@@
 which is what Salty uses to create an executable software implementation of the controller. 
-This Moore machine encodes all the states of the controller, including a single initial state. 
+This Moore machine encodes all the states of the controller. %, including a single initial state. 
 It also encodes the transition relation between states, i.e. the controller's next state given the current state 

### 2020-05-28 doc/2020-SEFM/paper.tex
Incorporating most recent paper comments

@@
 Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
-since the required computational complexity is relatively low, and several tools now exist for GR(1) synthesis.
+since the required computational complexity is relatively low, and several tools exist for synthesis from GR(1) specifications.
 However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
@@
 We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
-Toward this end we have modified Salty -- a tool that produces executable software implementations of controllers 
+Toward this end, we have modified Salty -- a tool that produces executable software implementations of controllers 
 from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
@@
 In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
-a subset of properties comprising GR(1) specifications, namely system initial and transition properties. 
-We also discuss lessons learned about how to best encode controllers in SPARK for verification, 
+a subset of properties comprising GR(1) specifications, namely system initial and system transition properties. 
+We also discuss lessons learned about how to best encode controllers synthesized from GR(1) specifications in SPARK for verification, 
 examples in which verification found unexpected controller behaviors, and 
@@
 An alternative is to use ``correct-by-construction'' synthesis approaches to automatically generate 
-the design of a system's control logic directly from specifications, which can 
-reduce both design time and the likelihood of errors 
+a system's control logic directly from specifications, which can 
+reduce both the amount of time needed for design and the likelihood of errors 
 \cite{alur2016compositional}, \cite{fainekos2009temporal}, \cite{guo2014cooperative}, \cite{kupermann2001synthesizing}. 
@@
 For inputs in the set $\inputs$ controlled by the environment 
-and outputs in the set $\outputs$ produced by the system, terms are:
+and outputs in the set $\outputs$ produced by the system, terms are defined as:
 
@@
 
+\vspace{0.5em}
+
 \noindent \begin{xtabular}{p{.045\columnwidth}p{.015\columnwidth}p{.860\columnwidth}}
@@
 \end{xtabular}
+
+\vspace{0.5em}
 
@@
 
-Several tools for GR(1) synthesis are available. 
+Several tools for synthesis from GR(1) specifications are available. 
 For example, RATSY %(Requirements Analysis Tool with Synthesis) 
@@
 including Python, C++, Java, and now SPARK. 
-Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,''
+Though all of these tools implement synthesis algorithms that are theoretically ``correct-by-construction,''
 tool implementation errors could still result in errors in synthesized controllers.
@@
 with a specification language and an associated verification toolset \cite{hoang2015spark}.
-SPARK translates code to the Why language for processing by the Why3 theorem prover, 
-and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
+% SPARK translates code to the Why language for processing by the Why3 theorem prover, 
+% and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
 Though SPARK aims to perform fully automated verification, manual adaptation of the source code is in general necessary. 
@@
 one would have to remove unsupported features such as functions with side-effects, aliased variable names, exception handlers, etc.  
-Formal verification with SPARK also requires supplying requirements at the level of the source code in the 
-form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by indvidual subprograms.
-There are different levels to which one can use SPARK to verify a program. 
+Formal verification of user-specified properties also requires manually writing specifications at the level of the source code in the 
+form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by individual subprograms.
+There are different levels to which one can use SPARK to verify a program \cite{moy2019climbing}. 
 Generally, the level of verification performed during development is incremental, going from lowest to highest. 
-Verification at the lowest level, ``stone level,'' is achieved when code is accepted by SPARK, 
+Verification at the lowest level, colloquially referred to as ``stone level,'' is achieved when code is accepted by SPARK, 
 since the SPARK language has stricter legality rules than Ada. 
 Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
-flow analysis checks for errors such as reads of unintialized data or violations of user-specified data flow contracts. 
+flow analysis checks for errors such as reads of uninitialized data or violations of user-specified data flow contracts. 
 Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
-e.g. division by zero and numeric overflow/underflow. 
+e.g. division by zero or numeric overflow/underflow. 
 Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
@@
 At the moment, the SPARK code generated by Salty proves at gold level.
-SPARK is able to prove the system initial and system transition property portions of the original GR(1) specification. 
+In particular, SPARK is able to prove the system initial and system transition portions the original GR(1) specification. 
 Proof of system transition properties in turn requires proving a type invariant on the underlying Moore machine representation of the controller; 
 this type invariant encodes that each state is reached by a specific set of input values and produces a specific set of output values. 
-The platinum level would be reached if the liveness properties were expressed and proved, 
+Platinum level would be reached if the liveness properties were expressed and proved, 
 which we leave for future work.
@@
 To evaluate whether verification of synthesized SPARK controllers is feasible and has utility, 
-we pulled in a large body of examples from various sources. 
+we collected a corpus of examples from various sources. 
 In \sectref{sec:vipExample}, we take a detailed look at one of these examples, a controller that coordinates the actions of 
@@
 In \sectref{sec:caseStudies}, we give an overview of results for the rest of the examples, with a focus on scalability. 
-In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to structure synthesized controllers for proof.  
+In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to best structure synthesized controllers for proof.  
 % examples in which SPARK revealed unexpected controller behaviors, and caveats related to GR(1) specifications. 
@@
 and the next set of input values. 
-Note that this transition relation is deterministic, i.e. there is at most one possible next state for each state and set of input values. 
+% Note that this transition relation is deterministic, i.e. there is at most one possible next state for each state and set of input values. 
+% Actually -- this is not true. There are examples where non-deterministic transitions are possible. In this case, Salty picks one arbitrarily.
 Every state then encodes a set of output values to be produced each time the controller transitions to that state. 
@@
 Controllers synthesized from GR(1) specifications have the additional property that for each state, 
-the set of input values that brings the controller into that state is the same, regardless of what the previous state was.
+there is a unique set of input values that brings the controller into that state, regardless of what the previous state was.
 % there is exactly one set of input values that brings the controller into that state.
 
-If SPARK is chosen as the target progamming language, Salty translates this textual Moore machine representation 
+If SPARK is chosen as the target programming language, Salty translates this textual Moore machine representation 
 of the controller to an implementation in SPARK. 
@@
 For the SPARK implementation, Salty therefore generates functions corresponding to $\varphi^e_i$ and $\varphi^e_t$,
-which are used as a precondition for the move procedure. 
+which are used in the precondition for the move procedure. 
 It also generates functions corresponding to $\varphi^s_i$ and $\varphi^s_t$, 
-which are used as a postcondition for the move procedure. 
+which are used in the postcondition for the move procedure. 
 If SPARK can verify that the move function satisfies the postcondition given the precondition, 
@@
 ``state to input mapping'' and ``state to output mapping'' functions that encode which input values 
-and output values are associated with each state; these are used to define an invariant for the controller.  
-This is essentially all this is required to prove system initial and system transition properties of synthesized controllers in SPARK.
+and output values are associated with each state; these are used to define a type invariant on the controller.  
+This is essentially all that is required to prove system initial and system transition properties of synthesized controllers in SPARK.
 
-To explain synthesized SPARK controllers and annotations in more detail, we consider a simple example involving a traffic light controller. 
+To understand synthesized SPARK controllers and annotations in more detail, consider a simple example involving a traffic light controller. 
 Let the controller's single input be $tick$. 
@@
 current input value(s), and current output value(s). 
-Its initial state is always the last possible state number. 
+The internal state number is always initialized to the last possible state number; 
+this ``controller initialization'' state encodes the status of the controller before any inputs are received, and the only transition out of this state are those allowed by $\varphi_i^e$.
 In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
@@
 Furthermore, recall that controllers synthesized from GR(1) specifications have the additional property that 
-there is exactly one set of input values that brought the controller into that state. 
-Therefore, \lstinline{Controller} includes a type invariant (lines 36-37) that captures this property, 
-where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} state whether the controller has the corresponding state number and input or output value(s), respectively. 
+there is exactly one set of input values that brings the controller into each state. 
+\lstinline{Controller} therefore includes a type invariant (lines 36-37) that captures this property, 
+where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} evaluate 
+whether the controller's state corresponds to the expected input and output value(s), respectively. 
 % They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
@@
 unless it is specified during compilation that they should be.
-Executing ghost code is often used for debugging, e.g. to check whether an unproved property is actually true. 
+Executing ghost code is often used for debugging, e.g. to check an unproved property through testing. 
 In this case, removing the ghost code can save significant memory, 
@@
 \sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
-since \lstinline{tick = False} brings the system into state 1 and $red$ is true and $yellow$ and $green$ are false in state 1.
+since \lstinline{tick = False} brings the system into state 1, and $red$ is true and $yellow$ and $green$ are false in state 1.
 
-The public function \lstinline{Is_Init} (line 8) checks whether the controller is in its initial state, i.e. no inputs have yet been received. 
+The public function \lstinline{Is_Init} (line 8) checks whether the controller is in its initialization state, i.e. no inputs have yet been received. 
 The public function \lstinline{Env_Init} (line 9) checks whether input(s) satisfy $\varphi_e^i$. 
@@
 It is implemented as an expression function for the same reason.  
-As pulled from the Salty specification, $\varphi^s_i = red \land \lnot yellow \land \lnot green$, so this function returns the value of
+From the Salty specification, $\varphi^s_i = red \land \lnot yellow \land \lnot green$, so this function returns the value of
 the expression \lstinline{S.red and not S.yellow and not S.green}.
@@
 \lstinline{Is_Init} and \lstinline{Env_Init} are used for proof but are also callable in functional code. 
-We chose to make these non-ghost functions for reasons related to the meaning of GR(1) specifications. 
+We chose to make these functions non-ghost functions for reasons related to the meaning of GR(1) specifications. 
 Recall that GR(1) specifications have the form $\varphi_e \limplies \varphi_s$. 
@@
 is satisfied regardless of whether the system produces output value(s) that satisfy $\varphi_s$. 
-In theory, the system in this case could produce arbitrary outputs and still satisfy the overall specification. 
+In theory, the system could then produce arbitrary outputs and still satisfy the overall specification. 
 In practice, we believe a user would generally want to know that the environment violated its specification, 
-so that the user could either choose the system output value(s) explicitly or fall back to some other error handling routine. 
+so that the user could either choose the system output value(s) explicitly or fall back to some other routine. 
 Therefore, a user needs to be able to check inputs with \lstinline{Env_Init} if \lstinline{Is_Init} returns true, 
@@
 For the same reasons as above, \lstinline{Sys_Trans} is a ghost function but \lstinline{Env_Trans} is not.
-Note that \lstinline{Env_Trans} has a precondition that the controller must not be in its initial state, 
+Note that \lstinline{Env_Trans} has a precondition that the controller must not be in its initialization state, 
 since $\varphi_e^t$ can depend on both the current and next set of input value(s). 
-This precondition is also necessary for \lstinline{Sys_Trans}, because in the initial state, the attributes of the \lstinline{Controller} are not initialized, and the \lstinline{Sys_Trans} function might read these values.
-As with \lstinline{Env_Init} and \lstinline{Sys_Init}, the logic for these functions is pulled from the Salty specification and 
+This precondition is similarly necessary for \lstinline{Sys_Trans}, since it can depend on both current and next input and output value(s). 
+As with \lstinline{Env_Init} and \lstinline{Sys_Init}, the logic for these functions is generated from the Salty specification and 
 implemented in the body (not shown), since they make use of input and output values stored in the \lstinline{Controller}, whose fields are private.
 
-The public procedure \lstinline{Move} (lines 18-23) transitions a controller based on its current state and 
+The public procedure \lstinline{Move} (lines 18-23) transitions the \lstinline{Controller} based on 
+its current internally stored values (i.e. state number and most recent input and output values) and 
 next set of input value(s), and it produces the next set of output value(s). 
-It has a precondition that if the controller is in its initial state, 
+It has a precondition that if \lstinline{Controller} is in its initialization state (i.e. it has not yet received any inputs), 
 inputs must satisfy $\varphi_e^i$; otherwise they must satisfy $\varphi_e^t$.
@@
 The \lstinline{others} keyword can be used to cover the set of all input conditions not covered in any explicit cases of the contract. 
-Note that for the left-hand side of each case (left of the \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
+Note that for the left-hand side of each case (left of \lstinline{=>}), variable names refer to values before evaluation of the subprogram; 
 for the right-hand side, they refer to values after evaluation.
 Therefore, the aspect \lstinline{Old} can be used on the right to reference the value of a variable before evaluation of the subprogram. 
-Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initial state, then after execution of \lstinline{Move}, 
-the first set of output value(s) produced should satisfy $\varphi_s^i$ and the controller should no longer be in its initial state; 
-if it is not in its initial state, then the output value(s) produced should satisfy $\varphi_s^t$, 
-evaluated based on the most recent input and output values stored in \lstinline{C'Old}, the next input value(s) just provided (in this case in \lstinline{tick}), 
-and the next output values just generated (in this case in the record \lstinline{System}). 
-This set of contract cases embodies our main proof goal, i.e. verification of system initial and transition properties from the original GR(1) specification for the controller.
+Combined with the previous precondition, \lstinline{Contract_Cases} asserts that if the controller is in its initialization state, then after execution of \lstinline{Move}, 
+the first set of output value(s) produced should satisfy $\varphi_s^i$ and the controller should no longer be in its initialization state.  
+If the controller is not in its initialization state, then the output value(s) produced should satisfy $\varphi_s^t$, which is
+evaluated based on the most recent input and output values stored in \lstinline{C'Old}, the next input value(s) just provided (in this case stored in \lstinline{tick}), 
+and the next output values just generated (in this case stored in the record \lstinline{System}). 
+This set of contract cases embodies our main proof goal, i.e. verification of system initial and transition properties from the original GR(1) specification.
 
@@
 allowed out of each state. 
-In any case, SPARK will prove that such cases are not reachable.
+In any case, SPARK will prove that such cases are not reachable if the preconditions of \lstinline{Move} are met, i.e. if the environment satisfies its specification.
 
@@
 In such cases, enumerations or integers can make both specifications and code more compact and easier to read and understand. 
-As a technical aside, during synthesis, enumerations and integers are translated to a bit vector representation along with the necessary 
-initial and transition specifications to encode properties such as values being mutually exclusive and rules for addition and subtraction over integers. 
+As a technical aside, during synthesis, enumerations and integers are translated to a bit vector representation along with additional  
+initial and transition specifications that encode properties inherent to these types, such as 
+values being mutually exclusive and rules for addition and subtraction over integers. 
 Once synthesis is complete, values are translated back to their original enumeration or integer representation.
@@
 In order to evaluate the feasibility and utility of our approach, we collected GR(1) specifications from a variety of sources. 
-This includes the Salty repository, which contains examples of GR(1) specifications for control of UAV teams performing different missions. 
+This includes the Salty repository, which contains examples of GR(1) specifications for control of teams of UAVs performing different missions. 
 One of these encodes the rules for a ``VIP Escort'' mission, 
-in which one UAV is designated as a ``very important person'' (VIP) that must always be ``escorted'' by a friendly surveillance UAV when it moves
-and protected from an ``enemy'' UAV. 
+in which one UAV is designated as a ``very important person'' (VIP) that must always be ``escorted'' by a friendly surveillance UAV when it moves, 
+and it must also be protected from an ``enemy'' UAV. 
 The VIP and surveillance UAVs are controlled by the system, while the enemy UAV is controlled by the environment.
@@
 or when $vTrack_1$ changes from false to true, OpenUxAS commands surveillance UAV 1 to fly next to the VIP. 
-The high-level controller synthesized by Salty makes some assumptions about the behaviors implemented by OpenUxAS, 
+The high-level controller synthesized by Salty makes some assumptions about the low-level behaviors implemented by OpenUxAS, 
 in this case, mainly that all UAVs move at the same speed and transition to new regions at the same time. 
@@
 %
-{\small
+\small
 \begin{align}
@@
 \end{align}
-}
-\noindent These express \eqnref{eq:vipEnvInit} that the enemy UAV starts in region 4; 
+%
+These express \eqnref{eq:vipEnvInit} that the enemy UAV starts in region 4, and regions 3 and 5 start as surveilled; 
 \eqnref{eq:srSet} that a region is considered to be surveilled after either one of the surveillance UAVs is in it; 
@@
 \eqnref{eq:vipEnvDisallowedRegions} that the enemy UAV cannot go to regions 1 or 2; and
-\eqnref{eq:vipEnvLiveness} that the enemy UAV must infinitely often go to regions 3, 4, and 5.
+\eqnref{eq:vipEnvLiveness} that the enemy UAV must infinitely often not be in each region 3, 4, and 5.
 
 GR(1) specifications for the system are:
-{\small
+\small
 \begin{align}
 \varphi_s^i  = & \;\; (loc_v = 2) \land (loc_{s1} = 3) \land (loc_{s2} = 5) \land \lnot vTrack_1 \land \lnot vTrack_2 \label{eq:vipSysInit} \\
-\varphi_s^t  =  & \quad\quad\quad\;\; \always \Big( \lnot(loc_v = \lnext loc_v) \limplies (\lnext vTrack_1 \lor \lnext vTrack_2) \Big)  \land \mbox{} \label{eq:vipNoMoveUnlessTracked} \\
- & \bigwedge_{i = \{1 \ldots 2\}} \always \Big(vTrack_i \limplies (sloc_i = loc_v)\Big) \land \mbox{} \label{eq:vipSysVtrack} \\
-& \bigwedge_{i = \{1 \ldots 5\}} \always \Big((loc_v = i) \limplies \lnot(loc_e = i)\Big) \land \mbox{} \label{eq:vipSysNoEloc} \\
+\varphi_s^t  =  & \quad\quad\quad\quad\; \always \Big( \lnot(loc_v = \lnext loc_v) \limplies (\lnext vTrack_1 \lor \lnext vTrack_2) \Big)  \land \mbox{} \label{eq:vipNoMoveUnlessTracked} \\
+ & \;  \bigwedge_{i = \{1 \ldots 2\}} \;\;   \always \Big(vTrack_i \limplies (sloc_i = loc_v)\Big) \land \mbox{} \label{eq:vipSysVtrack} \\
+& \; \bigwedge_{i = \{1 \ldots 5\}} \;\;   \always \Big((loc_v = i) \limplies \lnot(loc_e = i)\Big) \land \mbox{} \label{eq:vipSysNoEloc} \\
  & \bigwedge_{i = \{v, s1, s2\}} \always \Big(\big( \lnext(loc_i = 1) \limplies (loc_i = 1) \lor (loc_i = 2) \lor (loc_i = 3)   \big) \land \mbox{} \label{eq:vipSysTrans1} \\
@@
   & \quad\quad\quad\quad\quad\;\;\; \big(\lnext (loc_i = 4) \limplies (loc_i = 3) \lor (loc_i = 4) \lor (loc_i = 5)   \big) \land \mbox{} \label{eq:vipSysTrans4} \\
-  & \quad\quad\quad\quad\quad\;\;\; \big(\lnext (loc_i = 5) \limplies (loc_i = 3) \lor (loc_i = 4) \lor (loc_i = 5)  \big)  \Big) \land \mbox{} \label{eq:vipSysTrans5} \\
+  & \quad\quad\quad\quad\quad\;\;\; \big(\lnext (loc_i = 5) \limplies (loc_i = 3) \lor (loc_i = 4) \lor (loc_i = 5)  \big)  \Big)  \label{eq:vipSysTrans5} \\
 \varphi_s^l = & \;\; \always \eventually (loc_v = 1) \land \always \eventually (loc_v = 5). \label{eq:vipSysLiveness}
 \end{align}
-}
-\noindent These express \eqnref{eq:vipSysInit} that the VIP starts in region 2, surveillance UAV 1 in region 3, and surveillance UAV 2 in region 5; 
+%
+These express \eqnref{eq:vipSysInit} that the VIP starts in region 2, surveillance UAV 1 in region 3, and surveillance UAV 2 in region 5; 
 \eqnref{eq:vipNoMoveUnlessTracked} that the VIP does not change regions unless a surveillance UAV is tracking it; 
@@
 analysis with SPARK revealed a previously undetected problem with this example's specification. 
-As written, the specification is synthesizable and produces what appears at a glance to be a reasonable controller with 97 states. 
+As written, the specification is realizable and produces what appears at a glance to be a reasonable controller with 97 states. 
 In fact, we had previously run this example with OpenAMASE, OpenUxAS, and a Python controller synthesized by Salty. 
-However, we did not notice that 34 of the controller's 97 states do not actually have successors. 
-In SPARK, this resulted in empty case statements in the \lstinline{Move} procedure. 
-For example, for state 2 the case alternative statement is simply \lstinline{when 2 =>}, i.e. there are no other statements associated with this case. 
-Attempting to execute this case alternative would result in a runtime error, which SPARK detected. 
+However, we did not originally notice in the Python implementation that 34 of the controller's 97 states do not have successors. 
+Since no special logic was added to Salty to handle this situation, the generated SPARK code included empty case statement alternatives in the \lstinline{Move} procedure. 
+For example, in state 2 the case statement alternative is simply \lstinline{when 2 =>}, with no statements in the body. 
+This code failed to compile, since there is no fall-through behavior in case statements in SPARK; 
+explicit statements are expected for each case statement alternative. 
 In Python, this error went undetected; 
@@
 
-The reason for this has to do with a subtlety the semantics of GR(1) specifications. 
+This error is the result of a subtlety of the semantics of GR(1) specifications. 
 Recall that GR(1) specifications take the form $\varphi_e \limplies \varphi_s$. 
@@
 in which the environment takes its turn first and the system takes its turn second. 
-The issue here is that the environment is able to take transitions that will necessarily cause it to violate $\varphi_e^t$. 
+The issue here is that the environment is able to take transitions that will necessarily cause it to violate $\varphi_e^t$ in the next time step. 
 Note that $\varphi_e^t$ contains terms of the form $\always \lnot p$, specifically $\always \lnot (loc_e = i)$ for $i = \{1, 2\}$  \eqnref{eq:vipEnvDisallowedRegions}. 
@@
 The distinction is important. 
-In the former case, the environment can choose $p$ for the ``next'' time step, and this does not violate $\always \lnot p$ 
-in the ``current'' time step, since the new ``current'' time step is only constructed once the environment and system have both taken their turn. 
+If the environment chooses $p$ for the ``next'' time step, this does not violate $\always \lnot p$ in the ``current'' time step.
+%, since the new ``current'' time step is only constructed once the environment and system have both taken their turn.
 However, once the next state is reached, $p$ becomes the new ``current'' value, and $\always \lnot p$ 
@@
 Generally, specifications should follow the latter form $\always \lnext \lnot p$, 
-and this was indeed an error in these specifications, so we changed $\always \lnot (loc_e = 1) \land \always \lnot (loc_e = 2)$
+which prohibits the environment from choosing $p$ in the ``next'' time step. 
+This was indeed an error in these specifications, so we changed $\always \lnot (loc_e = 1) \land \always \lnot (loc_e = 2)$
  to $\always \lnext \lnot (loc_e = 1) \land \always \lnext \lnot (loc_e = 2)$ in \eqnref{eq:vipEnvDisallowedRegions}.
-However, we also modified Salty to \lstinline{raise Program_Error} for such cases and checked 
-that we were still able to prove $\varphi_e \limplies \varphi_s$ for these examples  
+However, we also modified Salty to \lstinline{raise Program_Error} in cases with no successors 
+and checked that we were still able to prove $\varphi_e \limplies \varphi_s$ 
 (since reaching these cases would require violating the precondition $\varphi_e$). 
-Such cases amount to instances in which the precondition on \lstinline{Move} is violated, which is why we allow the user to execute \lstinline{Env_Trans} 
+Such cases amount to instances in which the precondition on \lstinline{Move} is would have to be violated, which is why we allow the user to execute \lstinline{Env_Trans} 
 as discussed in \sectref{sec:implementation}. 
 We also plan to have Salty issue a warning when there are states with no successors, 
-as such cases are likely unintended by the user.
+since such cases are likely unintended by the user.
 
@@
 Anzu, LTLMoP, TuLiP, Slugs, and Salty, all of which make their examples publicly available for download. 
-These examples -- including GR(1) specifications in Salty format, synthesized SPARK packages, and SPARK analysis results -- are all available on the 
+GR(1) specifications in Salty format, synthesized SPARK packages, and SPARK analysis results for all of these examples are available on the 
 Salty GitHub repository, including the traffic light example of the previous section. 
@@
 (1) the generated code can be quite long, leading to memory errors when trying to prove the subprogram with SPARK, and
-(2) when proving the postcondition, since the solvers prove all cases at the same time and the number of cases grows 
+(2) since the solvers prove all case statement alternatives at the same time and the number of case statement alternatives grows 
 exponentially with the number of inputs, sometimes the solvers are not able to prove the postcondition.
@@
 % in a manner similar to a classic programmer splitting large subprograms into smaller ones. 
-This would allow SPARK to apply its modular analysis on several smaller procedures, thus enabling the proof on larger files. 
-We are currently investigating ways to split up the procedure that does not accidentally create more difficulties for the solvers.
+This would allow SPARK to apply modular analysis on several smaller procedures, thus enabling the proof on larger files. 
+We are currently investigating ways to split up the procedure that does not accidentally create more difficulties for the underlying solvers.
 
@@
 As a result, these controllers contained empty case statements in the \lstinline{Move} procedure. 
-We had previously run the Salty-synthesized controller in Python for the example in \sectref{sec:vipExample} and not noticed the error, 
+We had previously run the Salty-synthesized controller in Python for the example in \sectref{sec:vipExample} and had not noticed the error, 
 though it would have resulted in an unhandled runtime exception if one of the states without successors had been reached.
 
-A meta-analysis of SPARK timing results also revealed that other examples in our database that did not have any inputs, 
+A meta-analysis of SPARK timing results also revealed that other examples in our database did not have any inputs, 
 i.e. they amounted to synthesizing a system independent of an environment. 
@@
 and it was satisfying to be able to generate proofs using a single tool rather than having to use multiple tools to perform verification.
-Furthermore, the act of performing ``end-to-end'' verification with SPARK on such controllers was valuable because (1) it revealed implementation errors for certain 
-types of specifications that result in random runtime errors in other Salty target languages, and (2) it revealed cases in which controllers were inefficiently encoded, i.e.
+Furthermore, the act of performing ``end-to-end'' verification with SPARK on such controllers was valuable because (1) it revealed a type of specification error in some examples that would result in  runtime errors in other Salty target languages, and (2) it revealed cases in which controllers were inefficiently encoded, i.e.
 when there is no environment.
@@
 We are also interested in expressing and proving liveness properties. 
-Liveness properties will be more challenging to verify because they necessarily require reasoning about future states. 
-Verifying system liveness in SPARK will require something like encoding a lookahead and showing that certain states will inevitably be reached 
-when the environment satisfies its specification, which can also include liveness terms. 
+Liveness properties will be more challenging to verify because they necessarily require reasoning about future states beyond the ``next'' state. 
+Verifying system liveness in SPARK will require something like encoding a lookahead buffer and showing that certain states will inevitably be reached 
+when the environment satisfies its specification, which can itself also include liveness terms. 
 This is likely to result in complex first-order formulas with alternating quantification over time, which are 
-notoriously hard to handle for automated solvers, so discharging the resulting proof obligations may prove to be a challenge.
-To tackle this issue, collaboration with a model checker performing verification at the level of the input language might be appropriate.
+notoriously hard to handle in automated solvers, so discharging the resulting proof obligations may prove to be a challenge.
+To tackle this issue, collaboration with a model checker performing verification at the level of the input language might be more appropriate.
 

### 2020-05-29 doc/2020-SEFM/paper.tex
Small changes to the Salty in SPARK paper

@@
 however, at this level, the specifications only partially specify the desired behavior of the code.  
-Verification at the next level, ``platinum level,'' consists verifying a complete set of specifications.
+Verification at the next level, ``platinum level,'' consists of verifying a complete set of specifications.
 At the moment, the SPARK code generated by Salty proves at gold level.
@@
 The internal state number is always initialized to the last possible state number; 
-this ``controller initialization'' state encodes the status of the controller before any inputs are received, and the only transition out of this state are those allowed by $\varphi_i^e$.
+this ``controller initialization'' state encodes the status of the controller before any inputs are received, and the only transition(s) out of this state are those allowed by $\varphi_i^e$.
 In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
@@
 We have chosen to describe this particular example in detail because during the process of extending Salty to produce SPARK translations, 
-analysis with SPARK revealed a previously undetected problem with this example's specification. 
+we discovered a previously undetected problem with this example's specification. 
 As written, the specification is realizable and produces what appears at a glance to be a reasonable controller with 97 states. 
@@
 For example, in state 2 the case statement alternative is simply \lstinline{when 2 =>}, with no statements in the body. 
-This code failed to compile, since there is no fall-through behavior in case statements in SPARK; 
+This code failed to compile, since SPARK does not allow for fall-through behavior in case statements (nor does Ada); 
 explicit statements are expected for each case statement alternative. 
@@
 simply had an empty list of ``next state'' values. 
-If the system had attempted to transition states based on an empty list, the result would have been an unhandled runtime error.
+We briefly note that if we had encoded the controller using a map in SPARK, 
+the error would still have been detected through SPARK analysis rather than a syntactic check of the code; 
+the case statement encoding was chosen for efficiency reasons to be discussed in \sectref{sec:lessonsLearned}.
 
@@
 
-Analysis with SPARK did reveal some unexpected behaviors. 
-As discussed in \sectref{sec:vipExample}, there were two example controllers\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} 
+The process of encoding and analyzing controllers in SPARK did reveal some unexpected behaviors. 
+First, as discussed in \sectref{sec:vipExample}, there were two example controllers\footnote{Salty's vip\_orig.salt and Anzu's arbiter.salt} 
 with specifications that resulted in states with no successors. 
@@
 though it would have resulted in an unhandled runtime exception if one of the states without successors had been reached.  
-
-A meta-analysis of SPARK timing results also revealed that other examples in our database did not have any inputs, 
+Second, a meta-analysis of SPARK timing results also revealed that other examples in our database did not have any inputs, 
 i.e. they amounted to synthesizing a system independent of an environment. 

### 2020-07-10 doc/2020-SEFM/paper.tex
Paper edits

@@
 %
-\author{Author 1\inst{1} \and
-Author 2\inst{2} \and
-Author 3\inst{3}}
+% \author{Author 1\inst{1} \and
+% Author 2\inst{2} \and
+% Author 3\inst{3}}
 
-%\author{Laura Humphrey\inst{1} \and
-%James Hamil\inst{2} \and
-%Joffrey Huguet\inst{3}}
+\author{Laura R. Humphrey\inst{1} \and
+James Hamil\inst{2} \and
+Joffrey Huguet\inst{3}}
 %
 
-% \authorrunning{L. Humphrey et al.}
-\authorrunning{Author 1 et al.}
+\authorrunning{L. Humphrey et al.}
 
@@
 %
-\institute{Institute 1 \\
-\email{author1@email.com} \and
-Institute 2 \\
-\email{author2@email.com}  \and
-Institute 3 \\
- \email{author3@email.com}}
-%
-%\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
-%\email{laura.humphrey@us.af.mil} \and
-%LinQuest Corp., Beavercreek, OH 45431, USA \\
-%\email{james.hamil.ctr@us.af.mil}  \and
-%AdaCore, F-75009 Paris, France \\
-% \email{huguet@adacore.com}}
+\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
+\email{laura.humphrey@us.af.mil} \and
+LinQuest Corp., Beavercreek, OH 45431, USA \\
+\email{james.hamil.ctr@us.af.mil}  \and
+AdaCore, F-75009 Paris, France \\
+ \email{huguet@adacore.com}}
 %
@@
 
+GR(1) is a fragment of linear temporal logic. 
 A GR(1) specification $\varphi$ takes the form $\varphi = \varphi^e \limplies \varphi^s$, 
@@
 It is assumed that at each time step, the environment chooses an input from $\inputs$, then the system chooses an output from $\outputs$ in response. 
-The result of synthesis is a \emph{control protocol} or \emph{strategy} that can be expressed as a Moore machine. 
+Synthesis from GR(1) specifications can therefore be viewed as a two-player game between the system and the environment, 
+where the goal is for the system to satisfy $\varphi^s$ as long as the environment satisfies $\varphi^e$. 
+If this goal is achievable, then we say the specification is realizable, 
+and the result is a \emph{control protocol} or \emph{strategy} for the system that can be expressed as a Moore machine. 
 %Note that in the case of GR(1), synthesized Moore machines not only produce a single set of system output values in each state, 
@@
 Generally, the level of verification performed during development is incremental, going from lowest to highest. 
-Verification at the lowest level, colloquially referred to as ``stone level,'' is achieved when code is accepted by SPARK, 
+Verification at the lowest level, colloquially referred to as \emph{stone level}, is achieved when code is accepted by SPARK, 
 since the SPARK language has stricter legality rules than Ada. 
-Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
+Verification at the next level, \emph{bronze level}, is achieved when flow analysis returns with no error; 
 flow analysis checks for errors such as reads of uninitialized data or violations of user-specified data flow contracts. 
-Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
+Verification at the next level, \emph{silver level}, ensures that no runtime errors will be encountered when executing the program, 
 e.g. division by zero or numeric overflow/underflow. 
-Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
+Verification at the next level, \emph{gold level}, consists of verifying key user specifications, such as 
 type invariants and subprogram preconditions and postconditions; 
 however, at this level, the specifications only partially specify the desired behavior of the code.  
-Verification at the next level, ``platinum level,'' consists of verifying a complete set of specifications.
+Verification at the next level, \emph{platinum level}, consists of verifying a complete set of specifications.
 At the moment, the SPARK code generated by Salty proves at gold level.
@@
 In \sectref{sec:vipExample}, we take a detailed look at one of these examples, a controller that coordinates the actions of 
-a team of unmanned air vehicles (UAVs) performing an escort mission, 
+a team of unmanned air vehicles performing an escort mission, 
 and we describe how SPARK revealed an error in the controller's specification. 
@@
 current input value(s), and current output value(s). 
-The internal state number is always initialized to the last possible state number; 
+The internal state number is always initialized to the last possible state number, i.e. the largest value for \lstinline{State_Num}; 
 this ``controller initialization'' state encodes the status of the controller before any inputs are received, and the only transition(s) out of this state are those allowed by $\varphi_i^e$.
 In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
-When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
-Here, there is only one input, so it is not wrapped in a record. 
+When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} (not used here) or \lstinline{System} (lines 4-6), respectively. 
+There is only one input in this example, so it is not wrapped in a record, which is why there is no \lstinline{Environment} record in \figref{fig:trafficLightSpec}. 
 Since \lstinline{Controller} is a Moore machine, for each internal state, 
@@
 In order to evaluate the feasibility and utility of our approach, we collected GR(1) specifications from a variety of sources. 
-This includes the Salty repository, which contains examples of GR(1) specifications for control of teams of UAVs performing different missions. 
+This includes the Salty repository, which contains examples of GR(1) specifications 
+for control of teams of unmanned air vehicles (UAVs) performing different missions. 
 One of these encodes the rules for a ``VIP Escort'' mission, 
-in which one UAV is designated as a ``very important person'' (VIP) that must always be ``escorted'' by a friendly surveillance UAV when it moves, 
+in which one UAV is designated as a ``very important person'' (VIP) 
+that must always be ``escorted'' by a friendly surveillance UAV when it moves, 
 and it must also be protected from an ``enemy'' UAV. 
@@
 (since reaching these cases would require violating the precondition $\varphi_e$). 
-Such cases amount to instances in which the precondition on \lstinline{Move} is would have to be violated, which is why we allow the user to execute \lstinline{Env_Trans} 
+Such cases amount to instances in which the precondition on \lstinline{Move} would have to be violated, 
+which is why we allow the user to execute \lstinline{Env_Trans} 
 as discussed in \sectref{sec:implementation}. 
